---
title: "Convert the feature matrix to a long format"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
editor:
  markdown:
    wrap: 72
---

## Shiny Documents

In this Quarto document we create two helper functions the first takes the 
output from featureValues() function and transforms that data into the long 
form and the second function takes the data as stored in the long form 
and converts it back into a two-dimensional matrix.

```{r}
library(xcms)
library(MsExperiment)
library(Spectra)
library(tidyr)
library(dplyr)
```

```{r}
pth <- getwd()
pth
```

```{r}
load("data/qtof.RData")
```

```{r}
feature_Values <- featureValues(qtof)
head(feature_Values)
```
This function convert the feature matrix output to a long table format
```{r}
matrix_to_long <- function(feature_values) {
  feature_values %>%
    as.data.frame() %>%
    tibble::rownames_to_column("feature_id") %>%
    pivot_longer(
      cols = -feature_id,
      names_to = "dataOrigin",
      values_to = "value"
    )
}

# Here we call the function
long_format <- matrix_to_long(feature_Values)
head(long_format, 10)
nrow(long_format)
```

The function bellow takes the long form data and converts it back into
a two-dimensional matrix. 

```{r}
long_to_matrix <- function(long_df) {
  long_df %>%
    pivot_wider(
      names_from = dataOrigin,
      values_from = value
    ) %>%
    as.data.frame() %>%
    tibble::column_to_rownames("feature_id")
}


two_D_matrix_format <- long_to_matrix(long_format)
head(two_D_matrix_format) 
```

