---
title: "Selection of representative MS2 spectra per feature"
format: html
editor: 
  markdown: 
    wrap: 72
---

In this quarto document we filtered the QTOF neg of flax seed data. We
create the precursorPurity and PrecursorIntensity columns. we remove MS1
data, then we select the 10 % spectra with highest precursor purity, and
from those spectra we select the 10 % spectra with highest precursor
intensity and finally we select the spectrum with highest number of
fragments

```{r}
library(Spectra)
library(xcms)
library(MsExperiment)
library(S4Vectors)
library(BiocParallel)
register(SerialParam())

pth <- getwd()
pth

```

Load the qtof final object after preprocessing from
ftms_preprocessing.qmd file, and convert it to a spectra object

```{r}
load(file.path(pth, "data/qtof.RData"))
qtof_neg <- spectra(qtof)
```

We next calculate the precursor purity, by calling the
`precursorPurity()` function from the spectra package, for each MS2
spectrum as well as the precursor intensity (since that is not provided
in the original mzML files).

```{r}
precursorPurity <- function(x, tolerance = 0.3, ppm = 10) {
   # if (is.unsorted(rtime(x)))
    #    stop("Spectra with data origin ", dataOrigin(x[1L]),
     #        " are not increasingly sorted by retention time.")
    ms2_idx <- which(msLevel(x) == 2L)
    ms1_idx <- which(msLevel(x) == 1L)
    ratios <- rep(NA_real_, length(x))  
    pmzs <- precursorMz(x)
    for (i in ms2_idx) {
        ratio <- NA_real_
        j <- ms1_idx[ms1_idx < i]
        
        if (length(j)) {
            j <- j[length(j)]
            s1 <- filterMzRange(
                x[j], c(pmzs[i] + c(-1, 1) * (tolerance + pmzs[i] * ppm / 1e6)))
            
            pks <- peaksData(s1, c("mz", "intensity"))[[1L]]
            
            if (nrow(pks)) {
                intensities <- pks[, 2L]
                ratio <- max(intensities) / sum(intensities)
            }
        }
        
        ratios[i] <- ratio
    }
    
    ratios
}
```

```{r}
precPurity <- precursorPurity(qtof_neg)
spectraData(qtof_neg)$precursorIntensity <- NULL
precursorIntensity <- estimatePrecursorIntensity(qtof_neg, tolerance = 0.3, ppm = 10)
qtof_neg$precursorPurity <- precPurity
qtof_neg$precursorIntensity <- precursorIntensity
spectraData(qtof_neg)
```

We next extract all MS2 spectra that can be associated to the MS1
feature:

```{r}
qtof@spectra <- qtof_neg
feature_spectra <- featureSpectra(dropFilledChromPeaks(qtof), msLevel = 2, skipFilled = TRUE)
```

we verify if an MS2 Spectra is linked to more than one feature

```{r}
scanIndex.per.ori <- split(feature_spectra$scanIndex, feature_spectra$dataOrigin)
lapply(scanIndex.per.ori, function(x) any(duplicated(x)))
lapply(scanIndex.per.ori, function(x) x[duplicated(x)])
```

```{r}
ms2_id <- paste0(feature_spectra$dataOrigin, "_", feature_spectra$acquisitionNum)
ms2_table <- table(ms2_id)
ms2_mult <- names(ms2_table)[ms2_table > 1]
length(ms2_mult)
```

There is `r length(ms2_mult)` duplicated assignment

# Here we print the `spectraData()` of each MS2 that was assigned to different features in the same sample:

\*\* First file \*\*

```{r}
spectraData(feature_spectra)[spectraData(feature_spectra)$scanIndex == 226 & spectraData(feature_spectra)$dataOrigin == "C:\\Users\\amentag\\Desktop\\these\\These\\clones\\new_clones\\RDynLib_qtof\\data\\mzml\\S1Rb-neg-DDA2.mzML" ,]

```

```{r}
spectraData(feature_spectra)[spectraData(feature_spectra)$scanIndex == 1460 & spectraData(feature_spectra)$dataOrigin == "C:\\Users\\amentag\\Desktop\\these\\These\\clones\\new_clones\\RDynLib_qtof\\data\\mzml\\S1Rb-neg-DDA2.mzML",]
```

```{r}
spectraData(feature_spectra)[spectraData(feature_spectra)$scanIndex == 1462 & spectraData(feature_spectra)$dataOrigin == "C:\\Users\\amentag\\Desktop\\these\\These\\clones\\new_clones\\RDynLib_qtof\\data\\mzml\\S1Rb-neg-DDA2.mzML",]
```

```{r}
spectraData(feature_spectra)[spectraData(feature_spectra)$scanIndex == 1464 & spectraData(feature_spectra)$dataOrigin == "C:\\Users\\amentag\\Desktop\\these\\These\\clones\\new_clones\\RDynLib_qtof\\data\\mzml\\S1Rb-neg-DDA2.mzML",]
```

```{r}
spectraData(feature_spectra)[spectraData(feature_spectra)$scanIndex == 1478 & spectraData(feature_spectra)$dataOrigin == "C:\\Users\\amentag\\Desktop\\these\\These\\clones\\new_clones\\RDynLib_qtof\\data\\mzml\\S1Rb-neg-DDA2.mzML",]
```

```{r}
spectraData(feature_spectra)[spectraData(feature_spectra)$scanIndex ==  1474 & spectraData(feature_spectra)$dataOrigin == "C:\\Users\\amentag\\Desktop\\these\\These\\clones\\new_clones\\RDynLib_qtof\\data\\mzml\\S1Rb-neg-DDA2.mzML",]
```

Except the MS2 with th scan Index 226, all the other peaks were assigned
multiple times to the "FT15816" and "FT15816" features, both of those
features have a very close mzmed, that differs just by 0.001 and a large
difference in retention time, the first has 661.481 and the second
642.251 seconds.

\*\* Second file \*\*

```{r}
spectraData(feature_spectra)[spectraData(feature_spectra)$scanIndex ==  98 & spectraData(feature_spectra)$dataOrigin == "C:\\Users\\amentag\\Desktop\\these\\These\\clones\\new_clones\\RDynLib_qtof\\data\\mzml\\S1Ra-neg-DDA2.mzML",]
```

The two features has an mzmed difference of 0.033, and a large
difference in retention time, the first has 73.4488 and the second
59.6624 seconds.

\*\* third and final file \*\*

```{r}
spectraData(feature_spectra)[spectraData(feature_spectra)$scanIndex ==  1816 & spectraData(feature_spectra)$dataOrigin == "C:\\Users\\amentag\\Desktop\\these\\These\\clones\\new_clones\\RDynLib_qtof\\data\\mzml\\S1Rc-neg-DDA2.mzML",]
```

The two features has the same mzmed, and a large difference in retention
time, the first has 826.336 and the second 813.282 seconds.

# Some Visualization of the duplicated assignments

```{r}
ids2 <- c("FT15816", "FT15821")
a_chr2 <- featureChromatograms(qtof, features = ids2)

plot(a_chr2[1])
abline(v = 661.481 , col = "red", lty = 2)

plot(a_chr2[2])
abline(v = 642.251, col = "red", lty = 2)
```

```{r}
#' Get the ambiguous MS2 for current feature
fid_ms2_mult <- intersect(ms2_mult, ms2_id[feature_spectra$feature_id == "FT15816"])
#' Get the other features for the ambiguous MS2 of current feature
fid_other_features <- setdiff(feature_spectra$feature_id[ms2_id %in% fid_ms2_mult], "FT15821")
other_eic <- featureChromatograms(qtof, features = "FT15821")


fid_ms2 <- feature_spectra[feature_spectra$feature_id == "FT15816"]
fid_eic <- featureChromatograms(qtof, features = "FT15816")

par(mfrow = c(2, 1))
plot(fid_eic)
grid()
abline(v = rtime(fid_ms2), col = "blue")
abline(v = rtime(feature_spectra)[ms2_id %in% fid_ms2_mult], col = "red")

plot(other_eic)
grid()
abline(v = rtime(fid_ms2), col = "blue")
abline(v = rtime(feature_spectra)[ms2_id %in% fid_ms2_mult], col = "red")
```

all the 10 spectra from the first file are present in both the first and
the second features

```{r}
#' Getting the chromatographic peaks assigned to the first and to the
#' second feature
a <- chromPeaks(qtof)[featureDefinitions(qtof)["FT15816", "peakidx"][[1L]], ]
b <- chromPeaks(qtof)[featureDefinitions(qtof)[
                   fid_other_features, "peakidx"][[1L]], ]

par(mfrow = c(1, 1))
plot(1, 1, pch = NA, xlim = range(rbind(a, b)[, c("rtmin", "rtmax")]),
     ylim = range(rbind(a, b)[, c("mzmin", "mzmax")]))
grid()
rect(xleft = a[, "rtmin"], xright = a[, "rtmax"],
     ybottom = a[, "mzmin"], ytop = a[, "mzmax"], border = "#ff000080")
rect(xleft = b[, "rtmin"], xright = b[, "rtmax"],
     ybottom = b[, "mzmin"], ytop = b[, "mzmax"], border = "#0000ff80", lty = 2)
```

```{r}
a <- feature_spectra[ms2_id == ms2_mult[3L]]
print(spa <- spectraData(a))

a_chr <- featureChromatograms(qtof, features = a$feature_id)
par(mfrow = c(1, 2))
plot(a_chr[1])
abline(v = a$rtime, col = "red")
plot(a_chr[2])
abline(v = a$rtime, col = "red")

ms2_rt <- a$rtime
abs(ms2_rt - spa$feature_rtmed)
abs(ms2_rt[1] - spa$feature_rtmed[1])/abs(ms2_rt[2] - spa$feature_rtmed[2])
```

```{r}
a <- feature_spectra[ms2_id == ms2_mult[5L]]
print(spa <- spectraData(a))

a_chr <- featureChromatograms(qtof, features = a$feature_id)
par(mfrow = c(1, 2))
plot(a_chr[1])
abline(v = a$rtime, col = "red")
plot(a_chr[2])
abline(v = a$rtime, col = "red")

ms2_rt <- a$rtime
abs(ms2_rt - spa$feature_rtmed)
abs(ms2_rt[1] - spa$feature_rtmed[1])/abs(ms2_rt[2] - spa$feature_rtmed[2])
```

```{r}
# Extract peaks table
pks <- chromPeaks(qtof)

# Compute peakwidths
pw <- pks[, "rtmax"] - pks[, "rtmin"]


# Problematic IDs
problem_ids <- c("CP082844", "CP406275")

# Subset the peaks
problem_pks <- pks[rownames(pks) %in% problem_ids, , drop = FALSE]

# Add peakwidth column
problem_pks <- cbind(problem_pks, peakwidth = problem_pks[, "rtmax"] - problem_pks[, "rtmin"])

# Show them
problem_pks

# Position in distribution (percentile)
problem_percentiles <- sapply(problem_pks[, "peakwidth"], function(x) ecdf(pw)(x) * 100)

 data.frame(
   chrom_peak_id = rownames(problem_pks),
   peakwidth = problem_pks[, "peakwidth"],
   percentile = round(problem_percentiles, 2)
 )



```

# Resolving multiple assignments of MS2 spectra to features

We next evaluate the cases above to reduce/remove the multiple
assignments and keep only a single MS2-feature association. First, we
select the duplicated MS2 spectra by identifying those with the same
*dataOrigin* and *acquisitionNum* but different *feature_id* to detect
duplicates assigned to multiple features. From these duplicated MS2
spectra, we then check first (*case A*) if a feature has unique MS2
spectra elsewhere. If that is the case any ambiguous assignment is
removed and only the unambiguous mappings are kept for the feature.
Subsequently, the assignments are checked again and if there are still
duplicated assignments the distance of the MS2 to the respective
features is evaluated (*case B*): we keep the top 10% MS2-feature
assignments. All other assignments are dropped for the respective
feature.

```{r}
select_unique_MS2 <- function(ms2, qtof, threshold = 0.002) {

  sp <- spectraData(ms2)

  # Crée un identifiant unique pour chaque groupe (dataOrigin + scanIndex)
  sp$groups <- paste0(sp$dataOrigin, "_", sp$scanIndex)

  # Indices de chaque groupe
  group_idx <- split(seq_len(nrow(sp)), sp$groups)

  # Initialisation
  sp$is_ambiguous <- FALSE
  sp$is_ambiguous[sp$groups %in% names(group_idx)[lengths(group_idx) > 1]] <- TRUE
  final_feature <- vector("list", nrow(sp))
  amb_f <- vector("list", nrow(sp))

  for (gname in names(group_idx)) {
    gidx <- group_idx[[gname]]
    feature_ids <- unique(sp$feature_id[gidx])

    # Si un seul feature, on le garde et rien n'est ambigu
    if (length(feature_ids) == 1) {
      final_feature[gidx] <- rep(list(feature_ids), length(gidx))
      amb_f[gidx] <- rep(list(character(0)), length(gidx))
      next
    }

    # MS2 precursor m/z et RT
    ms2_mz <- precursorMz(ms2)[gidx[1]]
    ms2_rt <- rtime(ms2)[gidx[1]]

    # mz et rt des features
    mz_features <- featureDefinitions(qtof)[feature_ids, "mzmed"]
    rt_features <- featureDefinitions(qtof)[feature_ids, "rtmed"]

    # Filtre par m/z
    mz_diff <- abs(ms2_mz - mz_features)
    valid_features <- feature_ids[mz_diff <= threshold]

    # Si aucune feature valide, tout est ambigu
    if (length(valid_features) == 0) {
      final_feature[gidx] <- rep(list(character(0)), length(gidx))
      amb_f[gidx] <- rep(list(feature_ids), length(gidx))
      next
    }

    # Si une seule feature valide, on la garde
    if (length(valid_features) == 1) {
      keep_feat <- valid_features
      final_feature[gidx] <- rep(list(keep_feat), length(gidx))
      amb_f[gidx] <- rep(list(setdiff(feature_ids, keep_feat)), length(gidx))
      next
    }

    # Plusieurs features valides : on garde celle la plus proche en RT
    rt_diff <- abs(ms2_rt - rt_features[valid_features])
    min_idx <- which.min(rt_diff)
    keep_feat <- valid_features[min_idx]

    # Assigne la feature retenue et les ambiguës à tous les indices du groupe
    final_feature[gidx] <- rep(list(keep_feat), length(gidx))
    amb_f[gidx] <- rep(list(setdiff(feature_ids, keep_feat)), length(gidx))
  }

  # Mise à jour de l'objet MS2
  ms2$is_ambiguous <- sp$is_ambiguous
  ms2$feature_id <- sapply(final_feature, function(x) if(length(x) == 0) NA_character_ else x)
  ms2$ambiguous <- amb_f

  # COLLAPSE des MS2 dupliqués (on garde la première ligne par groupe)
  unique_groups <- !duplicated(paste0(ms2$dataOrigin, "_", ms2$scanIndex))
  ms2 <- ms2[unique_groups, ]

  # Remplir feature_id à partir des ambiguës si feature_id est NA
  missing_idx <- is.na(ms2$feature_id)
  ms2$feature_id[missing_idx] <- sapply(ms2$ambiguous[missing_idx], function(x) {
    if(length(x) == 0) return(NA_character_)
    x[1]  # On choisit le premier feature comme sélection
  })

  # Supprimer le feature sélectionné de la liste ambiguous
  ms2$ambiguous <- lapply(seq_along(ms2$ambiguous), function(i) {
    setdiff(ms2$ambiguous[[i]], ms2$feature_id[i])
  })

  ms2
}






qtof_filtered <- select_unique_MS2(feature_spectra, qtof, threshold = 0.002)
```

```{r}
ms2_id <- paste0(qtof_filtered$dataOrigin, "_", qtof_filtered$acquisitionNum)
ms2_table <- table(ms2_id)
ms2_mult <- names(ms2_table)[ms2_table > 1]
length(ms2_mult)
```

Now after applying select_unique_MS2() function we don't have multiple
assignment

Now we test on some peaks that were assigned to more than one feature

```{r}
spectraData(qtof_filtered)[spectraData(qtof_filtered)$scanIndex ==  98 & spectraData(qtof_filtered)$dataOrigin == "C:\\Users\\amentag\\Desktop\\these\\These\\clones\\new_clones\\RDynLib_qtof\\data\\mzml\\S1Ra-neg-DDA2.mzML",]
```

```{r}
spectraData(qtof_filtered)[spectraData(qtof_filtered)$scanIndex ==  1474 & spectraData(qtof_filtered)$dataOrigin == "C:\\Users\\amentag\\Desktop\\these\\These\\clones\\new_clones\\RDynLib_qtof\\data\\mzml\\S1Rb-neg-DDA2.mzML",]
```

```{r}
spectraData(qtof_filtered)[spectraData(qtof_filtered)$scanIndex ==  98 & spectraData(qtof_filtered)$dataOrigin == "C:\\Users\\amentag\\Desktop\\these\\These\\clones\\new_clones\\RDynLib_qtof\\data\\mzml\\S1Ra-neg-DDA2.mzML",]
```

the result shows that duplicated spectra are removed

The table below summarizes the feature to MS2 spectra information.

```{r, results = "asis"}
library(pander)
tmp <- data.frame(no_features = nrow(featureDefinitions(qtof)),
                  no_features_ms2 = length(unique(feature_spectra$feature_id)),
                  no_selected_ms2 = length(feature_spectra)
                  )
pandoc.table(
    tmp, style = "rmarkdown",
    caption = paste0("Total number of features, features for which an MS2 was",
                     " found and number of selected MS2 spectra."))
```

Only for very few features an MS2 spectrum was found/defined.

#Now we process the selected MS2 spectra to select the *best
representative* MS2 spectrum per feature:

Select the 10 % spectra with highest precursor purity

```{r}
library(dplyr)

# Extract the spectra data and fragment counts
purity_df <- as.data.frame(spectraData(qtof_filtered)) |>
  mutate(
    fragment_count = lengths(qtof_filtered),
    index = row_number()
  )

purity_top <- purity_df |>
    group_by(feature_id) |>
    arrange(desc(precursorPurity)) |>
    mutate(n_rows = n(),
           top_n = ceiling(n_rows * 0.10),
           row_number = row_number()) |>
    filter(row_number <= top_n) |>
    select(-n_rows, -top_n, -row_number)

```

Of these spectra select the 10 % spectra with highest precursor
intensity

```{r}
intensity_top <- purity_top |>
  group_by(feature_id) |>
  arrange(desc(precursorIntensity)) |>
  mutate(n_rows = n(),
         top_n = ceiling(n_rows * 0.10),
         row_number = row_number()) |>
  filter(row_number <= top_n) |>
  select(-n_rows, -top_n, -row_number)

```

Of these spectra select the spectrum with highest number of fragments

```{r}
final_selected <- intensity_top |>
  group_by(feature_id) |>
  filter(fragment_count == max(fragment_count, na.rm = TRUE)) |>
  slice_head(n = 1) # in case of tie

# Extract those spectra from original object
qtof_filtered <- qtof_filtered[final_selected$index]

# Check number of unique features retained
length(unique(spectraData(qtof_filtered)$feature_id))
spectraData(qtof_filtered)

```

```{r}
length(unique(unlist(qtof_filtered$feature_id)))

```

As a result we have
`r length(unique(spectraData(qtof_filtered)$feature_id))` unique feature

Now we save the object

```{r}
save(qtof_filtered, file = file.path(pth, "data", "qtof_filtered.RData"))
```

## Session information

```{r}
sessionInfo()
```
