---
title: "Selection of representative MS2 spectra per feature"
format: html
editor: 
  markdown: 
    wrap: 72
---

In this quarto document we filtered the QTOF neg of flax seed data. We
create the precursorPurity and PrecursorIntensity columns. we remove MS1
data, then we select the 10 % spectra with highest precursor purity, and
from those spectra we select the 10 % spectra with highest precursor
intensity and finally we select the spectrum with highest number of
fragments

```{r}
library(Spectra)
library(xcms)
library(MsExperiment)
library(S4Vectors)
library(BiocParallel)
register(SerialParam())

pth <- getwd()
pth

```

Load the qtof final object after preprocessing from
ftms_preprocessing.qmd file, and convert it to a spectra object

```{r}
load(file.path(pth, "data/qtof.RData"))
qtof_neg <- spectra(qtof)
```

We next calculate the precursor purity, by calling the
`precursorPurity()` function from the spectra package, for each MS2
spectrum as well as the precursor intensity (since that is not provided
in the original mzML files).

```{r}
precursorPurity <- function(x, tolerance = 0.3, ppm = 10) {
   # if (is.unsorted(rtime(x)))
    #    stop("Spectra with data origin ", dataOrigin(x[1L]),
     #        " are not increasingly sorted by retention time.")
    ms2_idx <- which(msLevel(x) == 2L)
    ms1_idx <- which(msLevel(x) == 1L)
    ratios <- rep(NA_real_, length(x))
    pmzs <- precursorMz(x)
    for (i in ms2_idx) {
        ratio <- NA_real_
        j <- ms1_idx[ms1_idx < i]

        if (length(j)) {
            j <- j[length(j)]
            s1 <- filterMzRange(
                x[j], c(pmzs[i] + c(-1, 1) * (tolerance + pmzs[i] * ppm / 1e6)))

            pks <- peaksData(s1, c("mz", "intensity"))[[1L]]

            if (nrow(pks)) {
                intensities <- pks[, 2L]
                ratio <- max(intensities) / sum(intensities)
            }
        }

        ratios[i] <- ratio
    }

    ratios
}


```

```{r}
precPurity <- precursorPurity(qtof_neg)
spectraData(qtof_neg)$precursorIntensity <- NULL
precursorIntensity <- estimatePrecursorIntensity(qtof_neg, tolerance = 0.3, ppm = 10)
qtof_neg$precursorPurity <- precPurity
qtof_neg$precursorIntensity <- precursorIntensity
spectraData(qtof_neg)
```

We next extract all MS2 spectra that can be associated to the MS1
feature:

```{r}
qtof@spectra <- qtof_neg
feature_spectra <- featureSpectra(dropFilledChromPeaks(qtof), msLevel = 2, skipFilled = TRUE)
spectraData(feature_spectra)
```

\
Before centroiding:

```{r}
peaksData(feature_spectra)[[1]]
```

Print the peaks data for the example feature "FT17578" before
centroiding

```{r}
peaksData(feature_spectra)[spectraData(feature_spectra)$feature_id == "FT17578", ][[1]]
```

Centroiding the MS2 data

```{r}
feature_spectra_centro <- Spectra::pickPeaks(feature_spectra, halfWindowSize = 5L, k = 1L,  weighted = TRUE)
```

After centroiding

```{r}
peaksData(feature_spectra_centro)[[1]]
```

Print the peaks data for the example feature "FT17578" after centroiding

```{r}
peaksData(feature_spectra_centro)[spectraData(feature_spectra_centro)$feature_id == "FT17578", ][[1]]

```

we verify if an MS2 Spectra is linked to more than one feature

```{r}
scanIndex.per.ori <- split(feature_spectra_centro$scanIndex, feature_spectra_centro$dataOrigin)
lapply(scanIndex.per.ori, function(x) any(duplicated(x)))
lapply(scanIndex.per.ori, function(x) x[duplicated(x)])

```

```{r}
ms2_id <- paste0(feature_spectra_centro$dataOrigin, "_", feature_spectra_centro$acquisitionNum)
ms2_table <- table(ms2_id)
ms2_mult <- names(ms2_table)[ms2_table > 1]
length(ms2_mult)
```

There is `r length(ms2_mult)` duplicated assignment

# In case of duplicated assignments: Resolving multiple assignments of MS2 spectra to features

We next evaluate the cases above to reduce/remove the multiple
assignments and keep only a single MS2-feature association. First, we
select the duplicated MS2 spectra by identifying those with the same
*dataOrigin* and *acquisitionNum* but different *feature_id* to detect
duplicates assigned to multiple features. From these duplicated MS2
spectra, we then check first (*case A*) if a feature has unique MS2
spectra elsewhere. If that is the case any ambiguous assignment is
removed and only the unambiguous mappings are kept for the feature.
Subsequently, the assignments are checked again and if there are still
duplicated assignments the distance of the MS2 to the respective
features is evaluated (*case B*): we keep the top 10% MS2-feature
assignments. All other assignments are dropped for the respective
feature.

```{r}
filter_unique_MS2 <- function(ms2, threshold = 0.002) {

  sp <- spectraData(ms2)
  #'group by dataOrigin and scanIndex to identify
  #'duplicated assignments in each dataOrigin
  sp$ms2_id <- paste0(sp$dataOrigin, "_", sp$scanIndex)
  sp$feature_id <- as.character(sp$feature_id)
  sp$ambiguous <- vector("list", nrow(sp))

  # Find duplicated MS2
  ms2_table <- table(sp$ms2_id)
  duplicated_ms2 <- names(ms2_table)[ms2_table > 1]
  # keep unique MS2 assignments directly
  keep_rows <- which(!sp$ms2_id %in% duplicated_ms2)  

  if (length(duplicated_ms2) > 0) {
    # Get indices of duplicated MS2
    dup_idx <- which(sp$ms2_id %in% duplicated_ms2)

    # Split by ms2_id
    dup_groups <- split(dup_idx, sp$ms2_id[dup_idx])

    for (ms2_name in names(dup_groups)) {
      idx <- dup_groups[[ms2_name]]
      feature_ids <- sp$feature_id[idx]

      ms2_mz <- precursorMz(ms2)[idx[1]]
      ms2_rt <- rtime(ms2)[idx[1]]

      mz_features <- sp$feature_mzmed[idx]
      rt_features <- sp$feature_rtmed[idx]

      # m/z filter
      valid <- feature_ids[abs(ms2_mz - mz_features) <= threshold]

      if (length(valid) == 0) {
        # Keep first row, all features in ambiguous
        keep_row <- idx[1]
        sp$ambiguous[keep_row] <- list(feature_ids)
        keep_rows <- c(keep_rows, keep_row)
        next
      }

      # pick closest by RT if multiple
      if (length(valid) > 1) {
        rt_diff <- abs(ms2_rt - rt_features[match(valid, feature_ids)])
        keep_feat <- valid[which.min(rt_diff)]
      } else {
        keep_feat <- valid
      }

      # Keep only row corresponding to selected feature
      keep_row <- idx[which(feature_ids == keep_feat)[1]]
      keep_rows <- c(keep_rows, keep_row)

      # Store all other original features as ambiguous
      sp$ambiguous[keep_row] <- list(setdiff(feature_ids, keep_feat))
    }
  }

  ms2_filtered <- ms2[keep_rows, ]
  sp_final <- spectraData(ms2_filtered)
  sp_final$ambiguous <- sp$ambiguous[keep_rows]
  spectraData(ms2_filtered) <- sp_final

  ms2_filtered
}





qtof_filtered <- filter_unique_MS2(feature_spectra_centro, threshold = 0.002)
```

Removing the 10% with lowest purity

```{r}
# extract the spectra data as a dataframe
purity_df <- as.data.frame(spectraData(feature_spectra_centro)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", collisionEnergy)  
  )


keep_idx <- purity_df %>%
  group_by(f) %>%
  arrange(precursorPurity) %>%
  mutate(
    n_rows = n(),
    cutoff = ceiling(n_rows * 0.10),   # 10% lowest
    row_number = row_number()
  ) %>%
  filter(row_number > cutoff) %>%
  pull(index)


ms2_qtof <- feature_spectra_centro[keep_idx]

```

test the results on the previous feature

```{r}
peaksData(ms2_qtof)[spectraData(ms2_qtof)$feature_id == "FT17578", ][[1]]
```

create a single spectrum for each feature per collision energy, keeping
only fragments that are present in 50% of spectra

```{r}
# grouping by feature_id and collision energy
f <- paste(ms2_qtof$feature_id, "|", ms2_qtof$collisionEnergy)
ms2_qtof <- combineSpectra(ms2_qtof, ppm = 5, peaks = "intersect", minProp = 0.5, 
    intensityFun = max, f = f, p = rep(1L, length(ms2_qtof)))
```

test the results on the previous feature

```{r}
peaksData(ms2_qtof)[spectraData(ms2_qtof)$feature_id == "FT17578", ][[1]]
```

grouping by feature_id for different collision energies

```{r}
# grouping by feature_id for different collision energies
f = ms2_qtof$feature_id
ms2_qtof <- combineSpectra(ms2_qtof, ppm = 5, peaks = "union",intensityFun = sum,
                                       f = f, p = rep(1L, length(ms2_qtof)))
```

```{r}
spectraData(ms2_qtof)$spectrum_type <- "merged_MSMS_all_energies"
```

test the results on the previous feature to visualize the final selected
peaks for this feature

```{r}
peaksData(ms2_qtof)[spectraData(ms2_qtof)$feature_id == "FT17578", ][[1]]
```

```{r}
length(unique(unlist(ms2_qtof$feature_id)))

length(unique(spectraData(ms2_qtof)$feature_id))
```

As a result we have `r length(unique(spectraData(ms2_qtof)$feature_id))`
unique feature

Now we save the object

```{r}
save(ms2_qtof, file = file.path(pth, "data", "ms2_qtof.RData"))
```

## Session information

```{r}
sessionInfo()
```
