---
title: "Selection of representative MS2 spectra per feature"
format: html
editor: 
  markdown: 
    wrap: 72
---

In this quarto document we filtered the QTOF neg of flax seed data. We
create the precursorPurity and PrecursorIntensity columns. we remove MS1
data, then we select the 10 % spectra with highest precursor purity, and
from those spectra we select the 10 % spectra with highest precursor
intensity and finally we select the spectrum with highest number of
fragments

```{r}
library(Spectra)
library(xcms)
library(MsExperiment)
library(S4Vectors)
library(BiocParallel)
library(Rdisop)
library(stringr)
register(SerialParam())

pth <- getwd()
pth

```

Load the qtof final object after preprocessing from
ftms_preprocessing.qmd file, and convert it to a spectra object

```{r}
load(file.path(pth, "data/qtof.RData"))
qtof_neg <- spectra(qtof)
```

We next calculate the precursor purity, by calling the
`precursorPurity()` function from the spectra package, for each MS2
spectrum as well as the precursor intensity (since that is not provided
in the original mzML files).

```{r}
precursorPurity <- function(x, tolerance = 0.3, ppm = 10) {
   # if (is.unsorted(rtime(x)))
    #    stop("Spectra with data origin ", dataOrigin(x[1L]),
     #        " are not increasingly sorted by retention time.")
    ms2_idx <- which(msLevel(x) == 2L)
    ms1_idx <- which(msLevel(x) == 1L)
    ratios <- rep(NA_real_, length(x))
    pmzs <- precursorMz(x)
    for (i in ms2_idx) {
        ratio <- NA_real_
        j <- ms1_idx[ms1_idx < i]

        if (length(j)) {
            j <- j[length(j)]
            s1 <- filterMzRange(
                x[j], c(pmzs[i] + c(-1, 1) * (tolerance + pmzs[i] * ppm / 1e6)))

            pks <- peaksData(s1, c("mz", "intensity"))[[1L]]

            if (nrow(pks)) {
                intensities <- pks[, 2L]
                ratio <- max(intensities) / sum(intensities)
            }
        }

        ratios[i] <- ratio
    }

    ratios
}


```

```{r}
precPurity <- precursorPurity(qtof_neg)
spectraData(qtof_neg)$precursorIntensity <- NULL
precursorIntensity <- estimatePrecursorIntensity(qtof_neg, tolerance = 0.3, ppm = 10)
qtof_neg$precursorPurity <- precPurity
qtof_neg$precursorIntensity <- precursorIntensity
spectraData(qtof_neg)
qtof@spectra <- qtof_neg
```

## Associate MS2 spectra to features
We next extract all MS2 spectra that can be associated to the MS1
feature:

```{r}

feature_spectra <- featureSpectra(dropFilledChromPeaks(qtof), msLevel = 2, skipFilled = TRUE)
spectraData(feature_spectra)
```


We focus on a well characterized compound, [M-H]- = C31H31O19, mono-isotopic mass 707.146502.

```{r}
precmz.707 <- spectraData(feature_spectra)[,"precursorMz"]>707.14 & spectraData(feature_spectra)[,"precursorMz"] < 707.15
precmzs.707 <- spectraData(feature_spectra[precmz.707,])$precursorMz
rts.707 <-spectraData(feature_spectra[precmz.707,])$feature_rtmed
hist(rts.707)
table(round(rts.707))
#there are three retention times
fts.707 <-spectraData(feature_spectra[precmz.707,])$feature_id
table(fts.707)
#there are three features with precursormass 707
idx <- which(rownames(featureDefinitions(qtof)) %in% fts.707)
featureDefinitions(qtof)[idx,]

mim <- 707.146502
ppm <- (precmzs.707 - mim) / mim * 1e6
ppm
hist(precmzs.707)
median(precmzs.707)
#is there a relation between precursorPurity and accuracy of precursorMz (ppm)?
purity.707 <- spectraData(feature_spectra[precmz.707,])$precursorPurity
plot(purity.707, ppm,
     xlab = "Precursor Purity",
     ylab = "Mass Accuracy (ppm)",
     main = "Relation between precursorPurity and ppm error")
#No correlation between precursor purity and mass accuracy
sum(purity.707<0.975)/length(purity.707)
```
\
Before centroiding:
We will look at peaks data for features FT10698, FT10697, and FT10688, for each of the three energy levels.
```{r}
# grouping by feature_id and collision energy
library(stringr)
filenames <- feature_spectra$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")
f <- paste(feature_spectra$feature_id, "|", dda)

feature_spectra_FT10698_dda1 <- feature_spectra[f == "FT10698 | DDA1",]
feature_spectra_FT10698_dda2 <- feature_spectra[f == "FT10698 | DDA2",]
feature_spectra_FT10698_dda3 <- feature_spectra[f == "FT10698 | DDA3",]
length(peaksData(feature_spectra_FT10698_dda1))
length(peaksData(feature_spectra_FT10698_dda2))
length(peaksData(feature_spectra_FT10698_dda3))

#Number of peaks in spectra for this feature with DDA1, DDA2, and DDA3, before centroiding:
dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda1), function(x) {
  dim(x)[1]
})

summary(counts)
dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda2), function(x) {
  dim(x)[1]
})
summary(counts)

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda3), function(x) {
  dim(x)[1]
})
summary(counts)

#about 1000 peaks in each of these spectra

```


##Centroiding the MS2 data
### HWS = 5; k = 1

```{r}
feature_spectra_centro <- Spectra::pickPeaks(feature_spectra, halfWindowSize = 5L, k = 1L)
#Number of peaks in spectra for this feature with DDA1, DDA2, and DDA3, after centroiding:
feature_spectra_FT10698_dda1 <- feature_spectra_centro[f == "FT10698 | DDA1",]
feature_spectra_FT10698_dda2 <- feature_spectra_centro[f == "FT10698 | DDA2",]
feature_spectra_FT10698_dda3 <- feature_spectra_centro[f == "FT10698 | DDA3",]

feature_spectra_FT10697_dda1 <- feature_spectra_centro[f == "FT10697 | DDA1",]
feature_spectra_FT10697_dda2 <- feature_spectra_centro[f == "FT10697 | DDA2",]
feature_spectra_FT10697_dda3 <- feature_spectra_centro[f == "FT10697 | DDA3",]

feature_spectra_FT10688_dda1 <- feature_spectra_centro[f == "FT10688 | DDA1",]
feature_spectra_FT10688_dda2 <- feature_spectra_centro[f == "FT10688 | DDA2",]
feature_spectra_FT10688_dda3 <- feature_spectra_centro[f == "FT10688 | DDA3",]

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda1), function(x) {
  dim(x)[1]
})

summary(counts)
dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda2), function(x) {
  dim(x)[1]
})
summary(counts)

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda3), function(x) {
  dim(x)[1]
})
summary(counts)

#Around 100 peaks after centroiding
```

After centroiding

```{r}
peaksData(feature_spectra_centro)[[1]]
```

We want to check the accuracy of the mz values for a number of known fragments.

```{r}
# Define known fragment formulas
formulas <- c(
  "C27H25O16", "C25H23O15", "C26H25O13", "C23H21O13", 
  "C21H19O12", "C15H9O7", "C27H27O16", "C25H21O14",
  "C29H27O17"
)

# Calculate monoisotopic masses
masses <- sapply(formulas, function(f) getMolecule(f)$exactmass)

# ppm error function
ppm_error <- function(theo_mass, exp_mz) {
  1e6 * (exp_mz - theo_mass) / theo_mass
}

# Function to process one spectrum (matrix of mz-intensity)
analyze_spectrum <- function(mz.data, formulas, masses, spec_id) {
  data.frame(
    Spectrum = spec_id,
    Formula = formulas,
    ExactMass = masses,
    Closest_mz = sapply(masses, function(m) {
      mz.data[which.min(abs(mz.data - m))]
    }),
    PPM = sapply(masses, function(m) {
      closest <- mz.data[which.min(abs(mz.data - m))]
      ppm_error(m, closest)
    })
  )
}

# Put all three spectrum sets of feature_spectra_FT10698 into a named list
feature_spectra_FT10698 <- list(
  FT10698dda1 = feature_spectra_FT10698_dda1,
  FT10698dda2 = feature_spectra_FT10698_dda2,
  FT10698dda3 = feature_spectra_FT10698_dda3,
  FT10697dda1 = feature_spectra_FT10697_dda1,
  FT10697dda2 = feature_spectra_FT10697_dda2,
  FT10697dda3 = feature_spectra_FT10697_dda3,
  FT10688dda1 = feature_spectra_FT10688_dda1,
  FT10688dda2 = feature_spectra_FT10688_dda2,
  FT10688dda3 = feature_spectra_FT10688_dda3
)

# Loop over spectra objects
for (set_name in names(feature_spectra_FT10698)) {
  cat("\n==== Results for", set_name, "====\n")
  
  spectra <- feature_spectra_FT10698[[set_name]]
  
  # Run across all spectra in this feature set
  all_results <- do.call(rbind, lapply(seq_along(spectra), function(i) {
    mz.data <- peaksData(spectra)[[i]][,1]  # first column = m/z
    analyze_spectrum(mz.data, formulas, masses, i)
  }))
  
  # Filter only ppm < 5
  filtered_results <- subset(all_results, abs(PPM) < 5)
  
  # Table 1: matches per spectrum
  cat("\nTable: Matches per spectrum\n")
  print(table(filtered_results$Spectrum))
  
  # Table 2: how many spectra have N matches
  cat("\nTable: Spectra count by number of matches\n")
  print(table(table(filtered_results$Spectrum)))
}
```

Conclusion:
FT10688 is probably another compound.
FT10697 and FT10698 seem to be two isomers that fragment in a similar way.
MS/MS spectra of FT10698 show most matches with the given fragment list:
Up to 7 matching peaks in spectrum 12 of FT10698 with dda2.
Most matches are created with DDA3 (highest collision energy level).

### HWS = 5; k = 2

```{r}
feature_spectra_centro <- Spectra::pickPeaks(feature_spectra, halfWindowSize = 5L, k = 2L)
#Number of peaks in spectra for this feature with DDA1, DDA2, and DDA3, after centroiding:
feature_spectra_FT10698_dda1 <- feature_spectra_centro[f == "FT10698 | DDA1",]
feature_spectra_FT10698_dda2 <- feature_spectra_centro[f == "FT10698 | DDA2",]
feature_spectra_FT10698_dda3 <- feature_spectra_centro[f == "FT10698 | DDA3",]

feature_spectra_FT10697_dda1 <- feature_spectra_centro[f == "FT10697 | DDA1",]
feature_spectra_FT10697_dda2 <- feature_spectra_centro[f == "FT10697 | DDA2",]
feature_spectra_FT10697_dda3 <- feature_spectra_centro[f == "FT10697 | DDA3",]

feature_spectra_FT10688_dda1 <- feature_spectra_centro[f == "FT10688 | DDA1",]
feature_spectra_FT10688_dda2 <- feature_spectra_centro[f == "FT10688 | DDA2",]
feature_spectra_FT10688_dda3 <- feature_spectra_centro[f == "FT10688 | DDA3",]

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda1), function(x) {
  dim(x)[1]
})

summary(counts)
dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda2), function(x) {
  dim(x)[1]
})
summary(counts)

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda3), function(x) {
  dim(x)[1]
})
summary(counts)

#Around 100 peaks after centroiding
```


We want to check the accuracy of the mz values for a number of known fragments.

```{r}

# Loop over spectra objects
for (set_name in names(feature_spectra_FT10698)) {
  cat("\n==== Results for", set_name, "====\n")
  
  spectra <- feature_spectra_FT10698[[set_name]]
  
  # Run across all spectra in this feature set
  all_results <- do.call(rbind, lapply(seq_along(spectra), function(i) {
    mz.data <- peaksData(spectra)[[i]][,1]  # first column = m/z
    analyze_spectrum(mz.data, formulas, masses, i)
  }))
  
  # Filter only ppm < 5
  filtered_results <- subset(all_results, abs(PPM) < 5)
  
  # Table 1: matches per spectrum
  cat("\nTable: Matches per spectrum\n")
  print(table(filtered_results$Spectrum))
  
  # Table 2: how many spectra have N matches
  cat("\nTable: Spectra count by number of matches\n")
  print(table(table(filtered_results$Spectrum)))
}
```
Conclusion:
HWS = 5; k = 2 gives a lower number of good matches than HWS = 5; k = 1.

### HWS = 4; k = 1

```{r}
feature_spectra_centro <- Spectra::pickPeaks(feature_spectra, halfWindowSize = 4L, k = 1L)
#Number of peaks in spectra for this feature with DDA1, DDA2, and DDA3, after centroiding:
feature_spectra_FT10698_dda1 <- feature_spectra_centro[f == "FT10698 | DDA1",]
feature_spectra_FT10698_dda2 <- feature_spectra_centro[f == "FT10698 | DDA2",]
feature_spectra_FT10698_dda3 <- feature_spectra_centro[f == "FT10698 | DDA3",]


dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda1), function(x) {
  dim(x)[1]
})

summary(counts)
dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda2), function(x) {
  dim(x)[1]
})
summary(counts)

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda3), function(x) {
  dim(x)[1]
})
summary(counts)

#Around 100 peaks after centroiding
```
```{r}

# Loop over spectra objects
for (set_name in names(feature_spectra_FT10698)) {
  cat("\n==== Results for", set_name, "====\n")
  
  spectra <- feature_spectra_FT10698[[set_name]]
  
  # Run across all spectra in this feature set
  all_results <- do.call(rbind, lapply(seq_along(spectra), function(i) {
    mz.data <- peaksData(spectra)[[i]][,1]  # first column = m/z
    analyze_spectrum(mz.data, formulas, masses, i)
  }))
  
  # Filter only ppm < 5
  filtered_results <- subset(all_results, abs(PPM) < 5)
  
  # Table 1: matches per spectrum
  cat("\nTable: Matches per spectrum\n")
  print(table(filtered_results$Spectrum))
  
  # Table 2: how many spectra have N matches
  cat("\nTable: Spectra count by number of matches\n")
  print(table(table(filtered_results$Spectrum)))
}
```
Conclusion: 
Pretty much the same as for HWS = 5, k = 1
We go on with the result of  `Spectra::pickPeaks(feature_spectra, halfWindowSize = 5L, k = 1L)`,
and we focus on spectrum [[4]] of `feature_spectra_FT10698_dda3` which had 6 matches with known fragments


Print the peaks data for the example feature "FT10698_dda3" after centroiding

```{r}
feature_spectra_FT10698_dda3 <- feature_spectra_centro[f == "FT10698 | DDA3",]
peaksData(feature_spectra_FT10698_dda3)[[4]]

mz.data <- peaksData(feature_spectra_FT10698_dda3)[[4]][,1]  # first column = m/z
matchfragments <- analyze_spectrum(mz.data, formulas, masses, i)
  
o <- order(abs(matchfragments$PPM))
matchfragments[o,]
```

## Verify if MS2 Spectra are linked to more than one feature

```{r}
scanIndex.per.ori <- split(feature_spectra_centro$scanIndex, feature_spectra_centro$dataOrigin)
lapply(scanIndex.per.ori, function(x) any(duplicated(x)))
lapply(scanIndex.per.ori, function(x) x[duplicated(x)])

```

```{r}
ms2_id <- paste0(feature_spectra_centro$dataOrigin, "_", feature_spectra_centro$acquisitionNum)
ms2_table <- table(ms2_id)
ms2_mult <- names(ms2_table)[ms2_table > 1]
length(ms2_mult)
```

There is `r length(ms2_mult)` duplicated assignment

## Precursor purity
Grouping by feature_id and collision energy
Removing per group the (10%) spectra with lowest precursor purity

```{r}
# extract the spectra data as a dataframe
library(d)
filenames <- feature_spectra$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")



purity_df <- as.data.frame(spectraData(feature_spectra_centro)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda) 
  )


keep_idx <- purity_df %>%
  group_by(f) %>%
  arrange(precursorPurity) %>%
  mutate(
    n_rows = n(),
    cutoff = ceiling(n_rows * 0.10),   # 10% lowest
    row_number = row_number()
  ) %>%
  filter(row_number > cutoff) %>%
  pull(index)
ms2_qtof_90pure <- feature_spectra_centro[keep_idx]

keep_idx <- purity_df %>%
  group_by(f) %>%
  arrange(precursorPurity) %>%
  mutate(
    n_rows = n(),
    cutoff = ceiling(n_rows * 0.20),   # 10% lowest
    row_number = row_number()
  ) %>%
  filter(row_number > cutoff) %>%
  pull(index)
ms2_qtof_80pure <- feature_spectra_centro[keep_idx]

keep_idx <- purity_df %>%
  group_by(f) %>%
  arrange(precursorPurity) %>%
  mutate(
    n_rows = n(),
    cutoff = ceiling(n_rows * 0.30),   # 10% lowest
    row_number = row_number()
  ) %>%
  filter(row_number > cutoff) %>%
  pull(index)
ms2_qtof_70pure <- feature_spectra_centro[keep_idx]

keep_idx <- purity_df %>%
  group_by(f) %>%
  arrange(precursorPurity) %>%
  mutate(
    n_rows = n(),
    cutoff = ceiling(n_rows * 0.50),   # 10% lowest
    row_number = row_number()
  ) %>%
  filter(row_number > cutoff) %>%
  pull(index)
ms2_qtof_50pure <- feature_spectra_centro[keep_idx]

keep_idx <- purity_df %>%
  group_by(f) %>%
  arrange(precursorPurity) %>%
  mutate(
    n_rows = n(),
    cutoff = ceiling(n_rows * 0.75),   # 10% lowest
    row_number = row_number()
  ) %>%
  filter(row_number > cutoff) %>%
  pull(index)
ms2_qtof_25pure <- feature_spectra_centro[keep_idx]
```

test the results on the previous feature

```{r}
#remove 75%


# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_25pure$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(ms2_qtof_25pure)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)
  )

#Get indices of spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# Run across only the filtered spectra (FT10698 | DDA3)
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  # extract m/z vector (first column)
  mz.data <- peaksData(ms2_qtof_80pure)[[spec_index]][,1]
  # readable label: "FT10698 | DDA3_1", "FT10698 | DDA3_2", ...
  spec_label <- paste0(spectra_df$f[spec_index], "_", j)
  analyze_spectrum(mz.data, formulas, masses, spec_label)
}))


# Filter only ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# Table 1: matches per spectrum
cat("\nTable: Matches per spectrum (spectrum label -> # matches)\n")
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  tbl1 <- table(filtered_results$Spectrum)
  print(tbl1)

  # Table 2: how many spectra have N matches
  cat("\nTable: Spectra count by number of matches (how many spectra had N matches)\n")
  print(table(tbl1))
}
#remove 50%


# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_50pure$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(ms2_qtof_50pure)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)
  )

#Get indices of spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# Run across only the filtered spectra (FT10698 | DDA3)
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  # extract m/z vector (first column)
  mz.data <- peaksData(ms2_qtof_80pure)[[spec_index]][,1]
  # readable label: "FT10698 | DDA3_1", "FT10698 | DDA3_2", ...
  spec_label <- paste0(spectra_df$f[spec_index], "_", j)
  analyze_spectrum(mz.data, formulas, masses, spec_label)
}))


# Filter only ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# Table 1: matches per spectrum
cat("\nTable: Matches per spectrum (spectrum label -> # matches)\n")
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  tbl1 <- table(filtered_results$Spectrum)
  print(tbl1)

  # Table 2: how many spectra have N matches
  cat("\nTable: Spectra count by number of matches (how many spectra had N matches)\n")
  print(table(tbl1))
}

#remove 30%


# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_70pure$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(ms2_qtof_70pure)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)
  )

#Get indices of spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# Run across only the filtered spectra (FT10698 | DDA3)
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  # extract m/z vector (first column)
  mz.data <- peaksData(ms2_qtof_80pure)[[spec_index]][,1]
  # readable label: "FT10698 | DDA3_1", "FT10698 | DDA3_2", ...
  spec_label <- paste0(spectra_df$f[spec_index], "_", j)
  analyze_spectrum(mz.data, formulas, masses, spec_label)
}))


# Filter only ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# Table 1: matches per spectrum
cat("\nTable: Matches per spectrum (spectrum label -> # matches)\n")
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  tbl1 <- table(filtered_results$Spectrum)
  print(tbl1)

  # Table 2: how many spectra have N matches
  cat("\nTable: Spectra count by number of matches (how many spectra had N matches)\n")
  print(table(tbl1))
}

#remove 20%


# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_80pure$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(ms2_qtof_80pure)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)
  )

#Get indices of spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# Run across only the filtered spectra (FT10698 | DDA3)
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  # extract m/z vector (first column)
  mz.data <- peaksData(ms2_qtof_80pure)[[spec_index]][,1]
  # readable label: "FT10698 | DDA3_1", "FT10698 | DDA3_2", ...
  spec_label <- paste0(spectra_df$f[spec_index], "_", j)
  analyze_spectrum(mz.data, formulas, masses, spec_label)
}))


# Filter only ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# Table 1: matches per spectrum
cat("\nTable: Matches per spectrum (spectrum label -> # matches)\n")
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  tbl1 <- table(filtered_results$Spectrum)
  print(tbl1)

  # Table 2: how many spectra have N matches
  cat("\nTable: Spectra count by number of matches (how many spectra had N matches)\n")
  print(table(tbl1))
}

#remove 10%


# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_90pure$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(ms2_qtof_90pure)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)
  )

#Get indices of spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# Run across only the filtered spectra (FT10698 | DDA3)
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  # extract m/z vector (first column)
  mz.data <- peaksData(ms2_qtof_90pure)[[spec_index]][,1]
  # readable label: "FT10698 | DDA3_1", "FT10698 | DDA3_2", ...
  spec_label <- paste0(spectra_df$f[spec_index], "_", j)
  analyze_spectrum(mz.data, formulas, masses, spec_label)
}))


# Filter only ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# Table 1: matches per spectrum
cat("\nTable: Matches per spectrum (spectrum label -> # matches)\n")
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  tbl1 <- table(filtered_results$Spectrum)
  print(tbl1)

  # Table 2: how many spectra have N matches
  cat("\nTable: Spectra count by number of matches (how many spectra had N matches)\n")
  print(table(tbl1))
}


```
### Conclusion PrecursorPurity
Removing the 10% or 20% low precursor purity spectra does not affect the best spectra.
The number of spectra with 6 matches remains unchanged (6).
The number of spectra with 5 matches reduces from 7 (originally, centroided) to 6 if 10% **or** 20% of the spectra with lowest precursor purity are discarded.
The number of poorer spectra, that had 4, 3, 2, or 1 matches with known fragments, decreased when the 10% spectra with lowest precursor purity were discarded and decreased even more when 20% was discarded.
Therefore, removing the 20% spectra with lowest precursor purity efficiently enriches the spectra in high quality spectra mz peaks that match known fragments.

### DIscard 20% spectra with lowest precursor purity

```{r}
# extract the spectra data as a dataframe
library(dplyr)
filenames <- feature_spectra$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")



purity_df <- as.data.frame(spectraData(feature_spectra_centro)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda) 
  )


keep_idx <- purity_df %>%
  group_by(f) %>%
  arrange(precursorPurity) %>%
  mutate(
    n_rows = n(),
    cutoff = ceiling(n_rows * 0.20),   # 20% lowest
    row_number = row_number()
  ) %>%
  filter(row_number > cutoff) %>%
  pull(index)
ms2_qtof <- feature_spectra_centro[keep_idx]

```

## Merge spectra within groups feature_id and collision energy
create a single spectrum for each feature per collision energy, keeping
only fragments that are present in 50% of spectra

```{r}
# grouping by feature_id and collision energy
f <- paste(ms2_qtof$feature_id, "|", ms2_qtof$collisionEnergy)
ms2_qtof <- combineSpectra(ms2_qtof, ppm = 5, peaks = "intersect", minProp = 0.5, 
    intensityFun = max, f = f, p = rep(1L, length(ms2_qtof)))
```

test the results on the previous feature

```{r}
peaksData(ms2_qtof)[spectraData(ms2_qtof)$feature_id == "FT17578", ][[1]]
```

grouping by feature_id for different collision energies

```{r}
# grouping by feature_id for different collision energies
f = ms2_qtof$feature_id
ms2_qtof <- combineSpectra(ms2_qtof, ppm = 5, peaks = "union",intensityFun = sum,
                                       f = f, p = rep(1L, length(ms2_qtof)))
```

```{r}
spectraData(ms2_qtof)$spectrum_type <- "merged_MSMS_all_energies"
```

test the results on the previous feature to visualize the final selected
peaks for this feature

```{r}
peaksData(ms2_qtof)[spectraData(ms2_qtof)$feature_id == "FT17578", ][[1]]
```

```{r}
length(unique(unlist(ms2_qtof$feature_id)))

length(unique(spectraData(ms2_qtof)$feature_id))
```

As a result we have `r length(unique(spectraData(ms2_qtof)$feature_id))`
unique feature

Now we save the object

```{r}
save(ms2_qtof, file = file.path(pth, "data", "ms2_qtof.RData"))
```

## Session information

```{r}
sessionInfo()
```
