---
title: "Flax Qtof negative to Sql format"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Rebecca Dauwe, Johannes Rainer"
---

## Shiny Documents

This Quarto document outlines the workflow for creating a SQL database from 
QTOF flax data. The dataset includes two types of spectra, identified by the
spectrum_type column: "merged_MSMS_all_energies" and "intersect_single_energy".

```{r}
library(Spectra)
library(xcms)
library(MsExperiment)
library(Spectra)
library(CompoundDb)
library(RSQLite)
library(S4Vectors) 
library(tidyr)
library(dplyr)
register(SerialParam())
setwd("C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib_qtof")
pth <- getwd()
pth
```

# Input data and directory structure
The working directory must be set to the root of the project. From this root directory, a subdirectory named `data/` is required. This directory contains **`Spectra` objects generated upstream** by the Quarto documents:

- `qtof_filtering.qmd`

The Project directory structure is as follows:

root/
├── qtof_filtering.qmd
├── qtofneg_flax_sql.qmd
└── data/
    ├── ms2_qtof_Comb.RData
    └── ms2_qtof_sep.RData
    └── qtof_flax_final.RData
    
The following data files must be present:

- **`data/ms2_qtof_sep.RData`**  
  contains single spectrum for each feature per collision energy of 
  the flax experiment.

- **`data/ms2_qtof_Comb.RData`**  
  Contains a `Spectra` after grouping by feature_id for different collision 
  energies of the flax experiment.
  
- **`data/qtof_flax_final.RData`**
  In this object we stored the both previous objects "ms2_qtof_sep.RData" and
  "ms2_qtof_Comb.RData".
  
All file paths used in this document are **relative to the project root directory**.

### Verify working directory

```{r}
# Verify working directory
pth <- getwd()
pth
```

##Here we load the qtof flax neg spectra object 
```{r}
load("data/qtof_flax_final.RData")
spectraData(qtof_flax_final)
```

##Create an empty sql database 
```{r}
con <- dbConnect(RSQLite::SQLite(), "data/QTOF_flax_neg.sqlite")
```

##Create table experiment to store experiment metadata
```{r}
#new_exp_id <- max(dbReadTable(con, "experiment")$expid, na.rm = TRUE) + 1
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS experiment (
    expid INTEGER PRIMARY KEY AUTOINCREMENT,  -- auto-incrementing ID
    date TEXT,
    user TEXT,
    machine TEXT,
    column TEXT,
    mstype TEXT,
    buffera TEXT,
    bufferb TEXT,
    gradient_time TEXT,
    source TEXT,
    mode TEXT,
    species TEXT,
    tissue TEXT,
    ce TEXT,
    meta TEXT
  );
")

```

##Insert Experiment data to Experiment table
```{r}
dbExecute(con, "
  INSERT INTO experiment 
    (date, user, machine, column, mstype, buffera, bufferb, gradient_time, source, mode, species, tissue, ce, meta)
  VALUES 
    ('2025-10-08', 'Ahlam', 'QTOF', NULL, 'QTOF', NULL, NULL, NULL, NULL, 'neg', NULL, 'seed', NULL, NULL);
")

```

Here we Print the Experiment table after creation and insertion
```{r}
dbGetQuery(con, "SELECT * FROM experiment;")

```

##Create table ms_compound to store compounds information
```{r}
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS ms_compound (
    compound_id TEXT PRIMARY KEY,
    expid INTEGER,
    nodename TEXT,
    retention_time REAL,
    mass_measured REAL,
    name TEXT,
    formula TEXT,
    exactmass REAL,
    ppm_deviation REAL,
    subsid INTEGER,
    conversion TEXT,
    wavelen REAL,
    smiles TEXT,
    isotope_ratio REAL,
    drift_time REAL,
    composition TEXT,
    inchi TEXT,
    inchikey TEXT,
    FOREIGN KEY (expid) REFERENCES experiment(expid)
  );
")


```

##Extract spectraData from the flax ftms negative object
```{r}
sp_data <- spectraData(qtof_flax_final)
head(sp_data[, c("feature_id", "feature_rtmed", "feature_mzmed")])

```

Insert data to the compound table
```{r}
cmp <- sp_data[, c("feature_id", "feature_rtmed", "feature_mzmed")] |>
  as.data.frame() |>
  unique()

n <- nrow(cmp)

# Get the last compound_id already in the table
last_id <- dbGetQuery(con, "SELECT compound_id FROM ms_compound ORDER BY compound_id DESC LIMIT 1;")$compound_id

#Determine the next numeric index
#start_num <- if(length(last_id) == 0) 1 else as.numeric(sub("DYL", "", last_id)) + 1

#  Generate the new compound_id vector
#compound_ids <- sprintf("DYL%07d", start_num:(start_num + n - 1))

# Retrieve all existing compound_id values from the table
compound_ids_raw <- dbReadTable(con, "ms_compound")$compound_id

# Convert to numeric to extract the maximum 
compound_ids <- suppressWarnings(as.numeric(as.character(compound_ids_raw)))
compound_ids <- compound_ids[!is.na(compound_ids)] 
# Find the next compound_id: 1 if empty, otherwise max + 1
start_id <- if (length(compound_ids) == 0) 1 else max(compound_ids) + 1

last_expid <- dbReadTable(con, "experiment")$expid
# Create the data frame with compound IDs
ms_compound_df <- data.frame(
  compound_id = as.character(seq(start_id, length.out = n)),
  expid = rep(last_expid, n),
  nodename = cmp$feature_id,
  retention_time = cmp$feature_rtmed,
  mass_measured = cmp$feature_mzmed,
  name = rep(NA_character_, n),
  formula = rep(NA_character_, n),
  exactmass = rep(NA_real_, n),
  ppm_deviation = rep(NA_real_, n),
  subsid = rep(NA_integer_, n),
  conversion = rep(NA_character_, n),
  wavelen = rep(NA_real_, n),
  smiles = rep(NA_character_, n),
  isotope_ratio = rep(NA_real_, n),
  drift_time = rep(NA_real_, n),
  composition = rep(NA_character_, n),
  inchi = rep(NA_character_, n),
  inchikey = rep(NA_character_, n),
  stringsAsFactors = FALSE
)


```

```{r}
#Insert ms_compound table 
dbWriteTable(con, "ms_compound", ms_compound_df, append = TRUE, row.names = FALSE)
```

```{r}
#Verification
dbGetQuery(con, "SELECT * FROM ms_compound LIMIT 10;")
```

#Create msms_spectrum table
```{r}
# Drop old table
dbExecute(con, "DROP TABLE IF EXISTS msms_spectrum;")

# Force all numeric columns in msms_spectrum to REAL
dbExecute(con, "
CREATE TABLE msms_spectrum (
  spectrum_id INTEGER PRIMARY KEY AUTOINCREMENT,
  compound_id TEXT,
  ms_level INTEGER,
  polarity INTEGER,
  spectrum_type TEXT,
  precursor_mz REAL,
  precursorIntensity REAL,
  precursorCharge INTEGER,
  collision_energy TEXT,
  isolationWindowLowerMz REAL,
  isolationWindowTargetMz REAL,
  isolationWindowUpperMz REAL,
  peaks_count INTEGER,
  totIonCurrent REAL,
  basePeakMZ REAL,
  basePeakIntensity REAL,
  ionisationEnergy REAL,
  lowMZ REAL,
  highMZ REAL,
  mergedScan INTEGER,
  mergedResultScanNum INTEGER,
  mergedResultStartScanNum INTEGER,
  mergedResultEndScanNum INTEGER,
  injectionTime REAL,
  filterString TEXT,
  spectrumId INTEGER,
  ionMobilityDriftTime REAL,
  scanWindowLowerLimit REAL,
  scanWindowUpperLimit REAL,
  electronBeamEnergy REAL,
  originalPrecursorMz REAL,
  precursorPurity REAL,
  chromPeakRT REAL,
  chromPeakMz REAL,
  chromPeakId TEXT,
  rtime REAL,
  scanIndex INTEGER,
  dataStorage TEXT,
  centroided INTEGER,
  smoothed INTEGER,
  instrument TEXT,
  splash TEXT,
  instrument_type TEXT,
  acquisitionNum INTEGER,
  precScanNum INTEGER,
  predicted REAL,
  dataOrigin TEXT,
  original_id TEXT,
  FOREIGN KEY (compound_id) REFERENCES ms_compound(compound_id)
);
")



```

## insert spectra data to the msms_spectrum table
```{r}
sp <- spectraData(qtof_flax_final)

getcol <- function(x) if (x %in% names(sp)) sp[[x]] else NA

msms_df <- data.frame(
  compound_id = NA,
  ms_level = getcol("msLevel"),
  polarity = getcol("polarity"),
  spectrum_type = getcol("spectrum_type"),
  precursor_mz = getcol("precursorMz"),
  precursorIntensity = getcol("precursorIntensity"),
  precursorCharge = getcol("precursorCharge"),  
  collision_energy = getcol("collisionEnergy"),
  isolationWindowLowerMz = getcol("isolationWindowLowerMz"),   
  isolationWindowTargetMz = getcol("isolationWindowTargetMz"),
  isolationWindowUpperMz = getcol("isolationWindowUpperMz"),
  peaks_count = getcol("peaksCount"),
  totIonCurrent = getcol("totIonCurrent"),
  basePeakMZ = getcol("basePeakMZ"),
  basePeakIntensity = getcol("basePeakIntensity"),
  ionisationEnergy = getcol("ionisationEnergy"),
  lowMZ = getcol("lowMZ"),
  highMZ = getcol("highMZ"),
  mergedScan = getcol("mergedScan"),
  mergedResultScanNum = getcol("mergedResultScanNum"),
  mergedResultStartScanNum = getcol("mergedResultStartScanNum"),
  mergedResultEndScanNum = getcol("mergedResultEndScanNum"),
  injectionTime = getcol("injectionTime"),
  filterString = getcol("filterString"),
  spectrumId = getcol("spectrumId"),
  ionMobilityDriftTime = getcol("ionMobilityDriftTime"),
  scanWindowLowerLimit = getcol("scanWindowLowerLimit"),
  scanWindowUpperLimit = getcol("scanWindowUpperLimit"),
  electronBeamEnergy = getcol("electronBeamEnergy"),
  originalPrecursorMz = getcol("originalPrecursorMz"),   
  precursorPurity = getcol("precursorPurity"),
  chromPeakRT = getcol("chromPeakRT"),
  chromPeakMz = getcol("chromPeakMz"),
  chromPeakId = getcol("chromPeakId"),
  rtime = getcol("rtime"),
  scanIndex = getcol("scanIndex"),
  dataStorage = getcol("dataStorage"),
  centroided = getcol("centroided"),
  smoothed = getcol("smoothed"),
  instrument = getcol("instrument"),
  instrument_type = getcol("instrument_type"),
  acquisitionNum = getcol("acquisitionNum"),
  precScanNum = getcol("precScanNum"),
  predicted = getcol("predicted"),
  dataOrigin = getcol("dataOrigin"),
  original_id = getcol("original_id"),
  stringsAsFactors = FALSE
)

msms_df$compound_id <- ms_compound_df$compound_id[match(sp$feature_id, ms_compound_df$nodename)]


```


```{r}
dbWriteTable(con, "msms_spectrum", msms_df, append = TRUE, row.names = FALSE)

```

Verification
```{r}
dbGetQuery(con, "SELECT * FROM msms_spectrum LIMIT 10;")
```

#Create msms_spectrum_peak
```{r}
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS msms_spectrum_peak (
    peak_id INTEGER PRIMARY KEY AUTOINCREMENT,
    spectrum_id INTEGER,
    mz REAL,
    intensity REAL,
    FOREIGN KEY (spectrum_id) REFERENCES msms_spectrum(spectrum_id)
  );
")


```

##insert each individual peak in a single row
```{r}
pd <- peaksData(qtof_flax_final)

# Flatten peaks list into a data frame
peaks_df <- do.call(
  rbind,
  lapply(seq_along(pd), function(i) {
    if (!is.null(pd[[i]]) && nrow(pd[[i]]) > 0) {
      data.frame(
        spectrum_id = i,  # corresponds to row order in msms_spectrum
        mz = pd[[i]][, 1],
        intensity = pd[[i]][, 2]
      )
    }
  })
)

head(peaks_df)

```


```{r}
dbWriteTable(con, "msms_spectrum_peak", peaks_df, append = TRUE, row.names = FALSE)

```


```{r}
dbGetQuery(con, "SELECT * FROM msms_spectrum_peak LIMIT 10;")
```

Create Synonym table
```{r}
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS Synonym (
    id INTEGER PRIMARY KEY
  )
")

```

#Create featue_matrix table

```{r}
dbExecute(con, "DROP TABLE IF EXISTS feature_matrix;")
```


```{r}
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS feature_matrix (
    nodename TEXT,
    dataOrigin TEXT,
    Value REAL,
    expid INTEGER,
    FOREIGN KEY (expid) REFERENCES experiment(expid)
  );
")

```

#Create the content of the feature matrix
```{r}
load("data/qtof.RData")
feature_Values <- featureValues(qtof)
head(feature_Values)
```

```{r}
matrix_to_long <- function(feature_values) {
  feature_values %>%
    as.data.frame() %>%
    tibble::rownames_to_column("feature_id") %>%
    pivot_longer(
      cols = -feature_id,
      names_to = "dataOrigin",
      values_to = "Value"
    )
}

# Here we call the function
long_format <- matrix_to_long(feature_Values)
head(long_format, 10)
nrow(long_format)
```

```{r}
# Get expid from experiment table
exp_id <- dbGetQuery(con, "SELECT expid FROM experiment LIMIT 1;")$expid

# Add expid column
long_format$expid <- exp_id


```

```{r}
names(long_format)[names(long_format) == "feature_id"] <- "nodename"

dbWriteTable(con, "feature_matrix", long_format, append = TRUE, row.names = FALSE)

```

```{r}
dbGetQuery(con, "Select * from feature_matrix LIMIT 10")
```

#Create metadata table
```{r}
md <- make_metadata(source = "Flax QTOF neg", url = NA_character_,
                    source_version = "1.0.0", source_date = "2025-10-08")
dbWriteTable(con, name = "metadata", md, row.names = FALSE)
```

#Here we print all the con tables
```{r}

dbListTables(con)
```

#Convert the data to compDb object
```{r}
cdb <- CompDb(con)
cdb
```

#Convert the compoundDb to a spectra object
```{r}
sp <- Spectra(cdb)
```

##print the spectraData of the resulting spectra object
```{r}
spectraData(sp, columns = c("msLevel", "compound_id", "nodename", "rtime", "retention_time", "spectrum_type"))
```

##print the spectraVariables of the resulting spectra object
```{r}
spectraVariables(sp)
```

