---
title: "Flax Qtof negative to Sql format"
format:
  html:
    toc: true
    self-contained: true
author: "Ahlam Mentag, Johannes Rainer"
---

## Shiny Documents

This Quarto document represents the folowing steps to create a Sql database for
Qtof Flax data

```{r}
library(Spectra)
library(xcms)
library(MsExperiment)
library(Spectra)
library(CompoundDb)
library(RSQLite)
library(S4Vectors) 
library(tidyr)
library(dplyr)
register(SerialParam())
setwd("C:/Users/amentag/Desktop/these/These/clones/new_clones/RDynLib_qtof")
pth <- getwd()
pth
```

```{r}
load("data/qtof_flax_final.RData")
spectraData(qtof_flax_final)
```

```{r}
con <- dbConnect(RSQLite::SQLite(), "data/QTOF_flax_neg.sqlite")
```

```{r}

dbExecute(con, "
  CREATE TABLE IF NOT EXISTS experiment (
    expid INTEGER PRIMARY KEY AUTOINCREMENT,  -- auto-incrementing ID
    date TEXT,
    user TEXT,
    machine TEXT,
    column TEXT,
    mstype TEXT,
    buffera TEXT,
    bufferb TEXT,
    gradient_time TEXT,
    source TEXT,
    mode TEXT,
    species TEXT,
    tissue TEXT,
    ce TEXT,
    meta TEXT
  );
")

```

```{r}
dbExecute(con, "
  INSERT INTO experiment 
    (date, user, machine, column, mstype, buffera, bufferb, gradient_time, source, mode, species, tissue, ce, meta)
  VALUES 
    ('2025-10-08', 'Ahlam', 'QTOF', NULL, 'QTOF', NULL, NULL, NULL, NULL, 'neg', NULL, 'seed', NULL, NULL);
")

```

```{r}
dbGetQuery(con, "SELECT * FROM experiment;")

```

Create table ms_compound
```{r}
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS ms_compound (
    compound_id TEXT PRIMARY KEY,
    expid INTEGER,
    node_name TEXT,
    retention_time REAL,
    mass_measured REAL,
    name TEXT,
    formula TEXT,
    exactmass REAL,
    ppm_deviation REAL,
    subsid INTEGER,
    conversion TEXT,
    wavelen REAL,
    smiles TEXT,
    isotope_ratio REAL,
    drift_time REAL,
    composition TEXT,
    inchi TEXT,
    inchikey TEXT,
    FOREIGN KEY (expid) REFERENCES experiment(expid)
  );
")


```

```{r}
sp_data <- spectraData(qtof_flax_final)
head(sp_data[, c("feature_id", "feature_rtmed", "feature_mzmed")])

```

```{r}
n <- nrow(sp_data)

# Get the last compound_id already in the table
last_id <- dbGetQuery(con, "SELECT compound_id FROM ms_compound ORDER BY compound_id DESC LIMIT 1;")$compound_id

#Determine the next numeric index
start_num <- if(length(last_id) == 0) 1 else as.numeric(sub("DYL", "", last_id)) + 1

#  Generate the new compound_id vector
compound_ids <- sprintf("DYL%07d", start_num:(start_num + n - 1))

# Create the data frame with compound IDs
ms_compound_df <- data.frame(
  compound_id = compound_ids,
  expid = rep(last_expid, n),
  node_name = sp_data$feature_id,
  retention_time = sp_data$feature_rtmed,
  mass_measured = sp_data$feature_mzmed,
  name = rep(NA_character_, n),
  formula = rep(NA_character_, n),
  exactmass = rep(NA_real_, n),
  ppm_deviation = rep(NA_real_, n),
  subsid = rep(NA_integer_, n),
  conversion = rep(NA_character_, n),
  wavelen = rep(NA_real_, n),
  smiles = rep(NA_character_, n),
  isotope_ratio = rep(NA_real_, n),
  drift_time = rep(NA_real_, n),
  composition = rep(NA_character_, n),
  inchi = rep(NA_character_, n),
  inchikey = rep(NA_character_, n),
  stringsAsFactors = FALSE
)


```

```{r}
#Insert ms_compound table 
dbWriteTable(con, "ms_compound", ms_compound_df, append = TRUE, row.names = FALSE)
```

```{r}
#Verification
dbGetQuery(con, "SELECT * FROM ms_compound LIMIT 10;")
```

Create msms_spectrum table
```{r}
# Drop old table
dbExecute(con, "DROP TABLE IF EXISTS msms_spectrum;")

# Force all numeric columns in msms_spectrum to REAL
dbExecute(con, "
CREATE TABLE msms_spectrum (
  spectrum_id INTEGER PRIMARY KEY AUTOINCREMENT,
  compound_id TEXT,
  ms_level INTEGER,
  polarity INTEGER,
  precursor_mz REAL,
  precursor_intensity REAL,
  precursor_charge INTEGER,
  collision_energy TEXT,  
  isolation_window_lower_mz REAL,
  isolation_window_target_mz REAL,
  isolation_window_upper_mz REAL,
  peaks_count INTEGER,
  tot_ion_current REAL,
  base_peak_mz REAL,
  base_peak_intensity REAL,
  ionisation_energy REAL,
  low_mz REAL,
  high_mz REAL,
  merged_scan INTEGER,
  merged_result_scan_num INTEGER,
  merged_result_start_scan_num INTEGER,
  merged_result_end_scan_num INTEGER,
  injection_time REAL,
  filter_string TEXT,
  spectrum_id_original INTEGER,
  ion_mobility_drift_time REAL,
  scan_window_lower_limit REAL,
  scan_window_upper_limit REAL,
  electron_beam_energy REAL,
  original_precursor_mz REAL,
  precursor_purity REAL,
  chrom_peak_rt REAL,
  chrom_peak_mz REAL,
  chrom_peak_id TEXT,
  rtime REAL,
  scan_index INTEGER,
  data_storage TEXT,
  centroided INTEGER,
  smoothed INTEGER,
  instrument TEXT,
  instrument_type TEXT,
  acquisition_num REAL,
  prec_scan_num INTEGER,
  predicted REAL,
  data_origin TEXT,
  splash TEXT,
  FOREIGN KEY (compound_id) REFERENCES ms_compound(compound_id)
);
")



```


```{r}
sp <- spectraData(qtof_flax_final)

getcol <- function(x) if (x %in% names(sp)) sp[[x]] else NA

msms_df <- data.frame(
  compound_id = NA,
  ms_level = getcol("msLevel"),
  polarity = getcol("polarity"),
  precursor_mz = getcol("precursorMz"),
  precursor_intensity = getcol("precursorIntensity"),
  precursor_charge = getcol("precursorCharge"),
  collision_energy = getcol("collisionEnergy"),
  isolation_window_lower_mz = getcol("isolationWindowLowerMz"),
  isolation_window_target_mz = getcol("isolationWindowTargetMz"),
  isolation_window_upper_mz = getcol("isolationWindowUpperMz"),
  peaks_count = getcol("peaksCount"),
  tot_ion_current = getcol("totIonCurrent"),
  base_peak_mz = getcol("basePeakMZ"),
  base_peak_intensity = getcol("basePeakIntensity"),
  ionisation_energy = getcol("ionisationEnergy"),
  low_mz = getcol("lowMZ"),
  high_mz = getcol("highMZ"),
  merged_scan = getcol("mergedScan"),
  merged_result_scan_num = getcol("mergedResultScanNum"),
  merged_result_start_scan_num = getcol("mergedResultStartScanNum"),
  merged_result_end_scan_num = getcol("mergedResultEndScanNum"),
  injection_time = getcol("injectionTime"),
  filter_string = getcol("filterString"),
  spectrum_id_original = getcol("spectrumId"),
  ion_mobility_drift_time = getcol("ionMobilityDriftTime"),
  scan_window_lower_limit = getcol("scanWindowLowerLimit"),
  scan_window_upper_limit = getcol("scanWindowUpperLimit"),
  electron_beam_energy = getcol("electronBeamEnergy"),
  original_precursor_mz = getcol("original_precursor_mz"),
  precursor_purity = getcol("precursorPurity"),
  chrom_peak_rt = getcol("chrom_peak_rt"),
  chrom_peak_mz = getcol("chrom_peak_mz"),
  chrom_peak_id = getcol("chrom_peak_id"),
  rtime = getcol("rtime"),
  scan_index = getcol("scanIndex"),
  data_storage = getcol("dataStorage"),
  centroided = getcol("centroided"),
  smoothed = getcol("smoothed"),
  instrument = getcol("instrument"),
  instrument_type = getcol("instrument_type"),
  acquisition_num = getcol("acquisitionNum"),
  prec_scan_num = getcol("precScanNum"),
  predicted = getcol("predicted"),
  data_origin = getcol("dataOrigin"),
  stringsAsFactors = FALSE
)
msms_df$compound_id <- ms_compound_df$compound_id[match(sp$feature_id, ms_compound_df$node_name)]

# Convert collision_energy to character (NA_character_ si vide)
msms_df$collision_energy <- as.character(msms_df$collision_energy)

# Les autres colonnes numériques restent numériques
msms_df$centroided <- as.integer(msms_df$centroided)
msms_df$smoothed <- as.integer(msms_df$smoothed)

numeric_cols <- c(
  "precursor_mz", "precursor_intensity",
  "isolation_window_lower_mz", "isolation_window_target_mz",
  "isolation_window_upper_mz", "tot_ion_current", "base_peak_mz",
  "base_peak_intensity", "ionisation_energy", "low_mz", "high_mz",
  "injection_time", "ion_mobility_drift_time",
  "scan_window_lower_limit", "scan_window_upper_limit",
  "electron_beam_energy", "original_precursor_mz",
  "precursor_purity", "chrom_peak_rt", "chrom_peak_mz",
  "rtime", "acquisition_num", "predicted"
)
msms_df[numeric_cols] <- lapply(msms_df[numeric_cols], as.numeric)


```


```{r}
dbWriteTable(con, "msms_spectrum", msms_df, append = TRUE, row.names = FALSE)

```

```{r}
dbGetQuery(con, "SELECT * FROM msms_spectrum LIMIT 10;")
```

Create msms_spectrum_peak
```{r}
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS msms_spectrum_peak (
    peak_id INTEGER PRIMARY KEY AUTOINCREMENT,
    spectrum_id INTEGER,
    mz REAL,
    intensity REAL,
    FOREIGN KEY (spectrum_id) REFERENCES msms_spectrum(spectrum_id)
  );
")


```


```{r}
pd <- peaksData(qtof_flax_final)

# Flatten peaks list into a data frame
peaks_df <- do.call(
  rbind,
  lapply(seq_along(pd), function(i) {
    if (!is.null(pd[[i]]) && nrow(pd[[i]]) > 0) {
      data.frame(
        spectrum_id = i,  # corresponds to row order in msms_spectrum
        mz = pd[[i]][, 1],
        intensity = pd[[i]][, 2]
      )
    }
  })
)

head(peaks_df)

```


```{r}
dbWriteTable(con, "msms_spectrum_peak", peaks_df, append = TRUE, row.names = FALSE)

```


```{r}
dbGetQuery(con, "SELECT * FROM msms_spectrum_peak LIMIT 10;")
```

Create Synonym table
```{r}
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS Synonym (
    id INTEGER PRIMARY KEY
  )
")

```

Create featue_matrix table
```{r}
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS feature_matrix (
    node_name TEXT,
    dataOrigin TEXT,
    Value REAL,
    expid INTEGER,
    PRIMARY KEY (node_name, dataOrigin, expid),
    FOREIGN KEY (expid) REFERENCES experiment(expid)
  );
")

```

Create the content of the feature matrix
```{r}
load("data/qtof.RData")
feature_Values <- featureValues(qtof)
head(feature_Values)
```

```{r}
matrix_to_long <- function(feature_values) {
  feature_values %>%
    as.data.frame() %>%
    tibble::rownames_to_column("feature_id") %>%
    pivot_longer(
      cols = -feature_id,
      names_to = "dataOrigin",
      values_to = "Value"
    )
}

# Here we call the function
long_format <- matrix_to_long(feature_Values)
head(long_format, 10)
nrow(long_format)
```

```{r}
# Get expid from experiment table
exp_id <- dbGetQuery(con, "SELECT expid FROM experiment LIMIT 1;")$expid

# Add expid column
long_format$expid <- exp_id


```

```{r}
names(long_format)[names(long_format) == "feature_id"] <- "node_name"

dbWriteTable(con, "feature_matrix", long_format, append = TRUE, row.names = FALSE)

```

```{r}
dbGetQuery(con, "Select * from feature_matrix LIMIT 10")
```

Create metadata table
```{r}
md <- make_metadata(source = "Flax QTOF neg", url = NA_character_,
                    source_version = "1.0.0", source_date = "2025-10-08")
dbWriteTable(con, name = "metadata", md, row.names = FALSE)
```

Here we print all the con tables
```{r}

dbListTables(con)
```

Convert the data to compDb object
```{r}
cdb <- CompDb(con)
cdb
```

Convert the compoundDb to a spectra object
```{r}
sp <- Spectra(cdb)
```

