---
title: "Selection of representative MS2 spectra per feature"
format: html
editor: 
  markdown: 
    wrap: 72
---
## Introdcution
In this quarto document we filtered the QTOF neg of flax seed data. We
create the precursorPurity and PrecursorIntensity columns. we remove MS1
data, then we select the 10 % spectra with highest precursor purity, and
from those spectra we select the 10 % spectra with highest precursor
intensity and finally we select the spectrum with highest number of
fragments

## Data upload

```{r, echo = FALSE}
library(Spectra)
library(xcms)
library(MsExperiment)
library(S4Vectors)
library(BiocParallel)
library(Rdisop)
library(stringr)
library(dplyr)
register(SerialParam())

pth <- getwd()
pth

```

Load the qtof final object after preprocessing from
ftms_preprocessing.qmd file, and convert it to a spectra object

```{r}
load(file.path(pth, "data/qtof.RData"))
qtof_neg <- spectra(qtof)
```

##Calculate Precursor purity

We next calculate the precursor purity, by calling the
`precursorPurity()` function from the spectra package, for each MS2
spectrum as well as the precursor intensity (since that is not provided
in the original mzML files).
### Function
```{r}
precursorPurity <- function(x, tolerance = 0.3, ppm = 10) {
   # if (is.unsorted(rtime(x)))
    #    stop("Spectra with data origin ", dataOrigin(x[1L]),
     #        " are not increasingly sorted by retention time.")
    ms2_idx <- which(msLevel(x) == 2L)
    ms1_idx <- which(msLevel(x) == 1L)
    ratios <- rep(NA_real_, length(x))
    pmzs <- precursorMz(x)
    for (i in ms2_idx) {
        ratio <- NA_real_
        j <- ms1_idx[ms1_idx < i]

        if (length(j)) {
            j <- j[length(j)]
            s1 <- filterMzRange(
                x[j], c(pmzs[i] + c(-1, 1) * (tolerance + pmzs[i] * ppm / 1e6)))

            pks <- peaksData(s1, c("mz", "intensity"))[[1L]]

            if (nrow(pks)) {
                intensities <- pks[, 2L]
                ratio <- max(intensities) / sum(intensities)
            }
        }

        ratios[i] <- ratio
    }

    ratios
}


```

### Calculation
```{r}
precPurity <- precursorPurity(qtof_neg)
spectraData(qtof_neg)$precursorIntensity <- NULL
precursorIntensity <- estimatePrecursorIntensity(qtof_neg, tolerance = 0.3, ppm = 10)
qtof_neg$precursorPurity <- precPurity
qtof_neg$precursorIntensity <- precursorIntensity
spectraData(qtof_neg)
qtof@spectra <- qtof_neg
```
### Save XcmsExperiment object with precursorPurity

```{r}
save(qtof, file = "qtof_precpur.RData")
```
### Load XcmsExperiment object with precursorPurity

```{r}
load(file = "qtof_precpur.RData")
```

## Associate MS2 spectra to features
We next extract all MS2 spectra that can be associated to the MS1
feature:

```{r}

feature_spectra <- featureSpectra(dropFilledChromPeaks(qtof), msLevel = 2, skipFilled = TRUE)
spectraData(feature_spectra)
```

## Explore previously characterized compound
We focus on a well characterized compound, [M-H]- = C31H31O19, mono-isotopic mass 707.146502.

```{r}
precmz.707 <- spectraData(feature_spectra)[,"precursorMz"]>707.14 & spectraData(feature_spectra)[,"precursorMz"] < 707.15
precmzs.707 <- spectraData(feature_spectra[precmz.707,])$precursorMz
rts.707 <-spectraData(feature_spectra[precmz.707,])$feature_rtmed
hist(rts.707)
table(round(rts.707))
#there are three retention times
fts.707 <-spectraData(feature_spectra[precmz.707,])$feature_id
table(fts.707)
#there are three features with precursormass 707
idx <- which(rownames(featureDefinitions(qtof)) %in% fts.707)
featureDefinitions(qtof)[idx,]

mim <- 707.146502
ppm <- (precmzs.707 - mim) / mim * 1e6
ppm
hist(precmzs.707)
median(precmzs.707)
#is there a relation between precursorPurity and accuracy of precursorMz (ppm)?
purity.707 <- spectraData(feature_spectra[precmz.707,])$precursorPurity
plot(purity.707, ppm,
     xlab = "Precursor Purity",
     ylab = "Mass Accuracy (ppm)",
     main = "Relation between precursorPurity and ppm error")
#No correlation between precursor purity and mass accuracy
sum(purity.707<0.970)/length(purity.707)
```
Really low purity, <0.97, represents about 5% of the spectra.

Before centroiding:
We will look at peaks data for features FT10698, FT10697, and FT10688, for each of the three energy levels.
```{r}
# grouping by feature_id and collision energy
filenames <- feature_spectra$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")
f <- paste(feature_spectra$feature_id, "|", dda)

feature_spectra_FT10698_dda1 <- feature_spectra[f == "FT10698 | DDA1",]
feature_spectra_FT10698_dda2 <- feature_spectra[f == "FT10698 | DDA2",]
feature_spectra_FT10698_dda3 <- feature_spectra[f == "FT10698 | DDA3",]
length(peaksData(feature_spectra_FT10698_dda1))
length(peaksData(feature_spectra_FT10698_dda2))
length(peaksData(feature_spectra_FT10698_dda3))

#Number of peaks in spectra for this feature with DDA1, DDA2, and DDA3, before centroiding:
dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda1), function(x) {
  dim(x)[1]
})

summary(counts)
dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda2), function(x) {
  dim(x)[1]
})
summary(counts)

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda3), function(x) {
  dim(x)[1]
})
summary(counts)


```
About 1000 peaks in each of these spectra, before centroiding

##Centroiding the MS2 data



### HWS = 5; k = 1

```{r}
feature_spectra_centro <- Spectra::pickPeaks(feature_spectra, halfWindowSize = 5L, k = 1L, threshold = 0.5)
#Number of peaks in spectra for this feature with DDA1, DDA2, and DDA3, after centroiding:
feature_spectra_FT10698_dda1 <- feature_spectra_centro[f == "FT10698 | DDA1",]
feature_spectra_FT10698_dda2 <- feature_spectra_centro[f == "FT10698 | DDA2",]
feature_spectra_FT10698_dda3 <- feature_spectra_centro[f == "FT10698 | DDA3",]

feature_spectra_FT10697_dda1 <- feature_spectra_centro[f == "FT10697 | DDA1",]
feature_spectra_FT10697_dda2 <- feature_spectra_centro[f == "FT10697 | DDA2",]
feature_spectra_FT10697_dda3 <- feature_spectra_centro[f == "FT10697 | DDA3",]

feature_spectra_FT10688_dda1 <- feature_spectra_centro[f == "FT10688 | DDA1",]
feature_spectra_FT10688_dda2 <- feature_spectra_centro[f == "FT10688 | DDA2",]
feature_spectra_FT10688_dda3 <- feature_spectra_centro[f == "FT10688 | DDA3",]

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda1), function(x) {
  dim(x)[1]
})

summary(counts)
dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda2), function(x) {
  dim(x)[1]
})
summary(counts)

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda3), function(x) {
  dim(x)[1]
})
summary(counts)


```
Around 100 peaks after centroiding.

After centroiding

We want to check the accuracy of the mz values for a number of known fragments.

```{r}
# Define known fragment formulas
formulas <- c(
  "C27H25O16", "C25H23O15", "C26H25O13", "C23H21O13", 
  "C21H19O12", "C15H9O7", "C27H27O16", "C25H21O14",
  "C29H27O17"
)

# Calculate monoisotopic masses
masses <- sapply(formulas, function(f) getMolecule(f)$exactmass)

# ppm error function
ppm_error <- function(theo_mass, exp_mz) {
  1e6 * (exp_mz - theo_mass) / theo_mass
}

# Function to process one spectrum (matrix of mz-intensity)
analyze_spectrum <- function(mz.data, formulas, masses, spec_id) {
  data.frame(
    Spectrum = spec_id,
    Formula = formulas,
    ExactMass = masses,
    Closest_mz = sapply(masses, function(m) {
      mz.data[which.min(abs(mz.data - m))]
    }),
    PPM = sapply(masses, function(m) {
      closest <- mz.data[which.min(abs(mz.data - m))]
      ppm_error(m, closest)
    })
  )
}

# Put all three spectrum sets of feature_spectra_FT10698 into a named list
feature_spectra_FT10698 <- list(
  FT10698dda1 = feature_spectra_FT10698_dda1,
  FT10698dda2 = feature_spectra_FT10698_dda2,
  FT10698dda3 = feature_spectra_FT10698_dda3,
  FT10697dda1 = feature_spectra_FT10697_dda1,
  FT10697dda2 = feature_spectra_FT10697_dda2,
  FT10697dda3 = feature_spectra_FT10697_dda3,
  FT10688dda1 = feature_spectra_FT10688_dda1,
  FT10688dda2 = feature_spectra_FT10688_dda2,
  FT10688dda3 = feature_spectra_FT10688_dda3
)

# Loop over spectra objects
for (set_name in names(feature_spectra_FT10698)) {
  cat("\n==== Results for", set_name, "====\n")
  
  spectra <- feature_spectra_FT10698[[set_name]]
  
  # Run across all spectra in this feature set
  all_results <- do.call(rbind, lapply(seq_along(spectra), function(i) {
    mz.data <- peaksData(spectra)[[i]][,1]  # first column = m/z
    analyze_spectrum(mz.data, formulas, masses, i)
  }))
  
  # Filter only ppm < 5
  filtered_results <- subset(all_results, abs(PPM) < 5)
  
  # Table 1: matches per spectrum
  cat("\nTable: Matches per spectrum\n")
  print(table(filtered_results$Spectrum))
  
  # Table 2: how many spectra have N matches
  cat("\nTable: Spectra count by number of matches\n")
  print(table(table(filtered_results$Spectrum)))
}
```

Conclusion:
FT10688 is probably another compound.
FT10697 and FT10698 seem to be two isomers that fragment in a similar way.
MS/MS spectra of FT10698 show most matches with the given fragment list:
Up to 7 matching peaks in spectrum 12 of FT10698 with dda2.
Most matches are created with DDA3 (highest collision energy level).

### HWS = 3; k = 3

```{r}
feature_spectra_centro <- Spectra::pickPeaks(feature_spectra, halfWindowSize = 3L, k = 3L, threshold = 0.5)
#Number of peaks in spectra for this feature with DDA1, DDA2, and DDA3, after centroiding:
feature_spectra_FT10698_dda1 <- feature_spectra_centro[f == "FT10698 | DDA1",]
feature_spectra_FT10698_dda2 <- feature_spectra_centro[f == "FT10698 | DDA2",]
feature_spectra_FT10698_dda3 <- feature_spectra_centro[f == "FT10698 | DDA3",]

feature_spectra_FT10697_dda1 <- feature_spectra_centro[f == "FT10697 | DDA1",]
feature_spectra_FT10697_dda2 <- feature_spectra_centro[f == "FT10697 | DDA2",]
feature_spectra_FT10697_dda3 <- feature_spectra_centro[f == "FT10697 | DDA3",]

feature_spectra_FT10688_dda1 <- feature_spectra_centro[f == "FT10688 | DDA1",]
feature_spectra_FT10688_dda2 <- feature_spectra_centro[f == "FT10688 | DDA2",]
feature_spectra_FT10688_dda3 <- feature_spectra_centro[f == "FT10688 | DDA3",]

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda1), function(x) {
  dim(x)[1]
})

summary(counts)
dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda2), function(x) {
  dim(x)[1]
})
summary(counts)

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda3), function(x) {
  dim(x)[1]
})
summary(counts)

#Around 130 peaks after centroiding
```

```{r}
# 1. Extract spectra info
filenames <- feature_spectra_centro$dataOrigin
dda <- stringr::str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(feature_spectra_centro)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)  # same format as "FT10698 | DDA3"
  )

# 2. Identify spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# 3. Run analysis only on these spectra
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  mz.data <- peaksData(feature_spectra_centro)[[spec_index]][,1]  # first column = m/z
  analyze_spectrum(mz.data, formulas, masses, j)
}))

# 4. Filter matches with ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# 5. Print tables
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  cat("\nMatches per spectrum (spectrum index -> # matches):\n")
  print(table(filtered_results$Spectrum))
  
  cat("\nSpectra count by number of matches (how many spectra had N matches):\n")
  print(table(table(filtered_results$Spectrum)))
}
```

### HWS = 5; k = 5

```{r}
feature_spectra_centro <- Spectra::pickPeaks(feature_spectra, halfWindowSize = 5L, k = 5L, threshold = 0.5)
#Number of peaks in spectra for this feature with DDA1, DDA2, and DDA3, after centroiding:
feature_spectra_FT10698_dda1 <- feature_spectra_centro[f == "FT10698 | DDA1",]
feature_spectra_FT10698_dda2 <- feature_spectra_centro[f == "FT10698 | DDA2",]
feature_spectra_FT10698_dda3 <- feature_spectra_centro[f == "FT10698 | DDA3",]

feature_spectra_FT10697_dda1 <- feature_spectra_centro[f == "FT10697 | DDA1",]
feature_spectra_FT10697_dda2 <- feature_spectra_centro[f == "FT10697 | DDA2",]
feature_spectra_FT10697_dda3 <- feature_spectra_centro[f == "FT10697 | DDA3",]

feature_spectra_FT10688_dda1 <- feature_spectra_centro[f == "FT10688 | DDA1",]
feature_spectra_FT10688_dda2 <- feature_spectra_centro[f == "FT10688 | DDA2",]
feature_spectra_FT10688_dda3 <- feature_spectra_centro[f == "FT10688 | DDA3",]

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda1), function(x) {
  dim(x)[1]
})

summary(counts)
dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda2), function(x) {
  dim(x)[1]
})
summary(counts)

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda3), function(x) {
  dim(x)[1]
})
summary(counts)

#Around 100 peaks after centroiding
```


We want to check the accuracy of the mz values for a number of known fragments.
```{r}
# 1. Extract spectra info
filenames <- feature_spectra_centro$dataOrigin
dda <- stringr::str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(feature_spectra_centro)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)  # same format as "FT10698 | DDA3"
  )

# 2. Identify spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# 3. Run analysis only on these spectra
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  mz.data <- peaksData(feature_spectra_centro)[[spec_index]][,1]  # first column = m/z
  analyze_spectrum(mz.data, formulas, masses, j)
}))

# 4. Filter matches with ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# 5. Print tables
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  cat("\nMatches per spectrum (spectrum index -> # matches):\n")
  print(table(filtered_results$Spectrum))
  
  cat("\nSpectra count by number of matches (how many spectra had N matches):\n")
  print(table(table(filtered_results$Spectrum)))
}
```

```{r}

# Loop over spectra objects
for (set_name in names(feature_spectra_FT10698)) {
  cat("\n==== Results for", set_name, "====\n")
  
  spectra <- feature_spectra_FT10698[[set_name]]
  
  # Run across all spectra in this feature set
  all_results <- do.call(rbind, lapply(seq_along(spectra), function(i) {
    mz.data <- peaksData(spectra)[[i]][,1]  # first column = m/z
    analyze_spectrum(mz.data, formulas, masses, i)
  }))
  
  # Filter only ppm < 5
  filtered_results <- subset(all_results, abs(PPM) < 5)
  
  # Table 1: matches per spectrum
  cat("\nTable: Matches per spectrum\n")
  print(table(filtered_results$Spectrum))
  
  # Table 2: how many spectra have N matches
  cat("\nTable: Spectra count by number of matches\n")
  print(table(table(filtered_results$Spectrum)))
}
```
Conclusion:
HWS = 5; k = 2 gives a lower number of good matches than HWS = 5; k = 1.

### HWS = 2; k = 2

```{r}
feature_spectra_centro <- Spectra::pickPeaks(feature_spectra, halfWindowSize = 2L, k = 2L, threshold = 0.5)
#Number of peaks in spectra for this feature with DDA1, DDA2, and DDA3, after centroiding:
feature_spectra_FT10698_dda1 <- feature_spectra_centro[f == "FT10698 | DDA1",]
feature_spectra_FT10698_dda2 <- feature_spectra_centro[f == "FT10698 | DDA2",]
feature_spectra_FT10698_dda3 <- feature_spectra_centro[f == "FT10698 | DDA3",]


dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda1), function(x) {
  dim(x)[1]
})

summary(counts)
dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda2), function(x) {
  dim(x)[1]
})
summary(counts)

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda3), function(x) {
  dim(x)[1]
})
summary(counts)

#Around 100 peaks after centroiding
```
About 200 peaks per spectrum with HWS=2, k=2


```{r}
# 1. Extract spectra info
filenames <- feature_spectra_centro$dataOrigin
dda <- stringr::str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(feature_spectra_centro)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)  # same format as "FT10698 | DDA3"
  )

# 2. Identify spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# 3. Run analysis only on these spectra
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  mz.data <- peaksData(feature_spectra_centro)[[spec_index]][,1]  # first column = m/z
  analyze_spectrum(mz.data, formulas, masses, j)
}))

# 4. Filter matches with ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# 5. Print tables
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  cat("\nMatches per spectrum (spectrum index -> # matches):\n")
  print(table(filtered_results$Spectrum))
  
  cat("\nSpectra count by number of matches (how many spectra had N matches):\n")
  print(table(table(filtered_results$Spectrum)))
}
```


```{r}

# Loop over spectra objects
for (set_name in names(feature_spectra_FT10698)) {
  cat("\n==== Results for", set_name, "====\n")
  
  spectra <- feature_spectra_FT10698[[set_name]]
  
  # Run across all spectra in this feature set
  all_results <- do.call(rbind, lapply(seq_along(spectra), function(i) {
    mz.data <- peaksData(spectra)[[i]][,1]  # first column = m/z
    analyze_spectrum(mz.data, formulas, masses, i)
  }))
  
  # Filter only ppm < 5
  filtered_results <- subset(all_results, abs(PPM) < 5)
  
  # Table 1: matches per spectrum
  cat("\nTable: Matches per spectrum\n")
  print(table(filtered_results$Spectrum))
  
  # Table 2: how many spectra have N matches
  cat("\nTable: Spectra count by number of matches\n")
  print(table(table(filtered_results$Spectrum)))
}
```

Print the peaks data for the example feature "FT10698_dda3" after centroiding with two different halfwindowSize settings

```{r}
# Run pickPeaks with two halfWindowSizes
spec_centro_hw2k2 <- Spectra::pickPeaks(feature_spectra,
                             halfWindowSize = 2L, k = 2L, threshold = 0.5)
spec_centro_hw3k3 <- Spectra::pickPeaks(feature_spectra,
                             halfWindowSize = 3L, k = 3L, threshold = 0.5)
spec_centro_hw5k5 <- Spectra::pickPeaks(feature_spectra,
                             halfWindowSize = 5L, k = 5L, threshold = 0.5)

# 1. Compare number of peaks per spectrum
summary(sapply(peaksData(spec_centro_hw2k2), nrow))
summary(sapply(peaksData(spec_centro_hw3k3), nrow))
summary(sapply(peaksData(spec_centro_hw5k5), nrow))

```

```{r}
feature_spectra_FT10698_dda3 <- spec_centro_hw2k2[f == "FT10698 | DDA3",]
spectraData(feature_spectra_FT10698_dda3[4])$dataOrigin
spectraData(feature_spectra_FT10698_dda3[4])$scanIndex
peaksData(feature_spectra_FT10698_dda3)[[4]]

mz.data <- peaksData(feature_spectra_FT10698_dda3)[[4]][,1]  # first column = m/z
matchfragments <- analyze_spectrum(mz.data, formulas, masses, i)
  
o <- order(abs(matchfragments$PPM))
matchfragments[o,]
```

```{r}
feature_spectra_FT10698_dda3 <- spec_centro_hw3k3[f == "FT10698 | DDA3",]
spectraData(feature_spectra_FT10698_dda3[4])$dataOrigin
spectraData(feature_spectra_FT10698_dda3[4])$scanIndex
peaksData(feature_spectra_FT10698_dda3)[[4]]

mz.data <- peaksData(feature_spectra_FT10698_dda3)[[4]][,1]  # first column = m/z
matchfragments <- analyze_spectrum(mz.data, formulas, masses, i)
  
o <- order(abs(matchfragments$PPM))
matchfragments[o,]
```


```{r}
feature_spectra_FT10698_dda3 <- spec_centro_hw5k5[f == "FT10698 | DDA3",]
spectraData(feature_spectra_FT10698_dda3[4])$dataOrigin
spectraData(feature_spectra_FT10698_dda3[4])$scanIndex
peaksData(feature_spectra_FT10698_dda3)[[4]]

mz.data <- peaksData(feature_spectra_FT10698_dda3)[[4]][,1]  # first column = m/z
matchfragments <- analyze_spectrum(mz.data, formulas, masses, i)
  
o <- order(abs(matchfragments$PPM))
matchfragments[o,]
```

### HWS = 2; k = 3

```{r}
feature_spectra_centro <- Spectra::pickPeaks(feature_spectra, halfWindowSize = 2L, k = 3L, threshold = 0.5)
#Number of peaks in spectra for this feature with DDA1, DDA2, and DDA3, after centroiding:
feature_spectra_FT10698_dda1 <- feature_spectra_centro[f == "FT10698 | DDA1",]
feature_spectra_FT10698_dda2 <- feature_spectra_centro[f == "FT10698 | DDA2",]
feature_spectra_FT10698_dda3 <- feature_spectra_centro[f == "FT10698 | DDA3",]


dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda1), function(x) {
  dim(x)[1]
})

summary(counts)
dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda2), function(x) {
  dim(x)[1]
})
summary(counts)

dim(peaksData(feature_spectra_FT10698_dda1)[[1]])[1]
counts <- sapply(peaksData(feature_spectra_FT10698_dda3), function(x) {
  dim(x)[1]
})
summary(counts)

#Around 100 peaks after centroiding
```
About 200 peaks per spectrum with HWS=2, k=4


```{r}
# 1. Extract spectra info
filenames <- feature_spectra_centro$dataOrigin
dda <- stringr::str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(feature_spectra_centro)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)  # same format as "FT10698 | DDA3"
  )

# 2. Identify spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# 3. Run analysis only on these spectra
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  mz.data <- peaksData(feature_spectra_centro)[[spec_index]][,1]  # first column = m/z
  analyze_spectrum(mz.data, formulas, masses, j)
}))

# 4. Filter matches with ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# 5. Print tables
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  cat("\nMatches per spectrum (spectrum index -> # matches):\n")
  print(table(filtered_results$Spectrum))
  
  cat("\nSpectra count by number of matches (how many spectra had N matches):\n")
  print(table(table(filtered_results$Spectrum)))
}
```


```{r}

# Loop over spectra objects
for (set_name in names(feature_spectra_FT10698)) {
  cat("\n==== Results for", set_name, "====\n")
  
  spectra <- feature_spectra_FT10698[[set_name]]
  
  # Run across all spectra in this feature set
  all_results <- do.call(rbind, lapply(seq_along(spectra), function(i) {
    mz.data <- peaksData(spectra)[[i]][,1]  # first column = m/z
    analyze_spectrum(mz.data, formulas, masses, i)
  }))
  
  # Filter only ppm < 5
  filtered_results <- subset(all_results, abs(PPM) < 5)
  
  # Table 1: matches per spectrum
  cat("\nTable: Matches per spectrum\n")
  print(table(filtered_results$Spectrum))
  
  # Table 2: how many spectra have N matches
  cat("\nTable: Spectra count by number of matches\n")
  print(table(table(filtered_results$Spectrum)))
}
```

Print the peaks data for the example feature "FT10698_dda3" after centroiding with two different halfwindowSize settings

```{r}
# Run pickPeaks with two halfWindowSizes
spec_centro_hw2k2 <- Spectra::pickPeaks(feature_spectra,
                             halfWindowSize = 2L, k = 2L, threshold = 0.5)
spec_centro_hw3k3 <- Spectra::pickPeaks(feature_spectra,
                             halfWindowSize = 3L, k = 3L, threshold = 0.5)
spec_centro_hw5k5 <- Spectra::pickPeaks(feature_spectra,
                             halfWindowSize = 5L, k = 5L, threshold = 0.5)

# 1. Compare number of peaks per spectrum
summary(sapply(peaksData(spec_centro_hw2k2), nrow))
summary(sapply(peaksData(spec_centro_hw3k3), nrow))
summary(sapply(peaksData(spec_centro_hw5k5), nrow))

```

```{r}
feature_spectra_FT10698_dda3 <- spec_centro_hw2k2[f == "FT10698 | DDA3",]
spectraData(feature_spectra_FT10698_dda3[4])$dataOrigin
spectraData(feature_spectra_FT10698_dda3[4])$scanIndex
peaksData(feature_spectra_FT10698_dda3)[[4]]

mz.data <- peaksData(feature_spectra_FT10698_dda3)[[4]][,1]  # first column = m/z
matchfragments <- analyze_spectrum(mz.data, formulas, masses, i)
  
o <- order(abs(matchfragments$PPM))
matchfragments[o,]
```

```{r}
feature_spectra_FT10698_dda3 <- spec_centro_hw3k3[f == "FT10698 | DDA3",]
spectraData(feature_spectra_FT10698_dda3[4])$dataOrigin
spectraData(feature_spectra_FT10698_dda3[4])$scanIndex
peaksData(feature_spectra_FT10698_dda3)[[4]]

mz.data <- peaksData(feature_spectra_FT10698_dda3)[[4]][,1]  # first column = m/z
matchfragments <- analyze_spectrum(mz.data, formulas, masses, i)
  
o <- order(abs(matchfragments$PPM))
matchfragments[o,]
```


```{r}
feature_spectra_FT10698_dda3 <- spec_centro_hw5k5[f == "FT10698 | DDA3",]
spectraData(feature_spectra_FT10698_dda3[4])$dataOrigin
spectraData(feature_spectra_FT10698_dda3[4])$scanIndex
peaksData(feature_spectra_FT10698_dda3)[[4]]

mz.data <- peaksData(feature_spectra_FT10698_dda3)[[4]][,1]  # first column = m/z
matchfragments <- analyze_spectrum(mz.data, formulas, masses, i)
  
o <- order(abs(matchfragments$PPM))
matchfragments[o,]
```
### Conclusion
The median number of peaks per spectrum is 144 for `(halfWindowSize = 2L, k = 2L)`, 102 for `(halfWindowSize = 3L, k = 3L)`, and 65 for `(halfWindowSize = 5L, k = 5L)`.

Most matches are found in **spectra from FT10698, with collision energy  DDA3**:
The  maximum number of matches with known fragments found in spectra for that feature and collision energy, was 8.
The number of spectra with 8 matches was 15 for `(halfWindowSize = 2L, k = 2L)`, 12 for `(halfWindowSize = 2L, k = 3L)` or `(halfWindowSize = 3L, k = 3L)`, 7 for `(halfWindowSize = 5L, k = 5L)`.
The number of spectra with >= 7 matches was 31 for `(halfWindowSize = 2L, k = 2L)`, 29 for `(halfWindowSize = 3L, k = 3L)`, and 26 for `(halfWindowSize = 5L, k = 5L)`.

So, centroiding with `(halfWindowSize = 2L, k = 2L)` results in the highest number of high quality spectra.

**spectrum S1Ra-neg-DDA3.mzM - scandindex 1295 (FT10698, with collision energy  DDA3):**
The setting (k = 1) vs. (hws = 2, k = 2, threshold = 0.5) vs. (hws = 3, k = 3, threshold = 0.5) did not alter the mz values of the known fragments. 
With (hws = 5, k = 5, threshold = 0.5), one mz (theoretically 607.1299) went totally off.
The number of peaks in the spectrum changed with the parameters:
hws5, k5: 118 peaks
hws3, k3: 192 peaks
hws2, k2: 262 peaks

We continue with `halfWindowSize = 2L, k = 2L, threshold = 0.5)`, which resulted in the largest number of high quality spectra, and hope that we can remove noise peaks in further steps.


### Final centroiding
```{r}
feature_spectra_centro <- Spectra::pickPeaks(feature_spectra, halfWindowSize = 2L, k = 2L, threshold = 0.5)

```
## Verify if MS2 Spectra are linked to more than one feature

```{r}
scanIndex.per.ori <- split(feature_spectra_centro$scanIndex, feature_spectra_centro$dataOrigin)


```

```{r}
ms2_id <- paste0(feature_spectra_centro$dataOrigin, "_", feature_spectra_centro$acquisitionNum)
ms2_table <- table(ms2_id)
ms2_mult <- names(ms2_table)[ms2_table > 1]
length(ms2_mult)
```

There is `r length(ms2_mult)` duplicated assignment

## Precursor purity
### Removing a fraction of the spectra based on precursor purity
Grouping by feature_id and collision energy
Removing per group the 5%, 10%, 20%, 30%, 50%, 75% spectra with lowest precursor purity.
Evaluate based on number of matches with known peaks.

```{r}
# extract the spectra data as a dataframe
filenames <- feature_spectra_centro$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")



purity_df <- as.data.frame(spectraData(feature_spectra_centro)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda) 
  )

# 5% lowest
keep_idx <- purity_df %>%
  group_by(f) %>%
  arrange(precursorPurity) %>%
  mutate(
    n_rows = n(),
    cutoff = ceiling(n_rows * 0.05),   
    row_number = row_number()
  ) %>%
  filter(row_number > cutoff) %>%
  pull(index)
ms2_qtof_95pure <- feature_spectra_centro[keep_idx]

# 10% lowest
keep_idx <- purity_df %>%
  group_by(f) %>%
  arrange(precursorPurity) %>%
  mutate(
    n_rows = n(),
    cutoff = ceiling(n_rows * 0.10),   
    row_number = row_number()
  ) %>%
  filter(row_number > cutoff) %>%
  pull(index)
ms2_qtof_90pure <- feature_spectra_centro[keep_idx]

# 20% lowest
keep_idx <- purity_df %>%
  group_by(f) %>%
  arrange(precursorPurity) %>%
  mutate(
    n_rows = n(),
    cutoff = ceiling(n_rows * 0.20),   
    row_number = row_number()
  ) %>%
  filter(row_number > cutoff) %>%
  pull(index)
ms2_qtof_80pure <- feature_spectra_centro[keep_idx]

# 30% lowest
keep_idx <- purity_df %>%
  group_by(f) %>%
  arrange(precursorPurity) %>%
  mutate(
    n_rows = n(),
    cutoff = ceiling(n_rows * 0.30),  
    row_number = row_number()
  ) %>%
  filter(row_number > cutoff) %>%
  pull(index)
ms2_qtof_70pure <- feature_spectra_centro[keep_idx]

# 50% lowest
keep_idx <- purity_df %>%
  group_by(f) %>%
  arrange(precursorPurity) %>%
  mutate(
    n_rows = n(),
    cutoff = ceiling(n_rows * 0.50),  
    row_number = row_number()
  ) %>%
  filter(row_number > cutoff) %>%
  pull(index)
ms2_qtof_50pure <- feature_spectra_centro[keep_idx]

# 75% lowest
keep_idx <- purity_df %>%
  group_by(f) %>%
  arrange(precursorPurity) %>%
  mutate(
    n_rows = n(),
    cutoff = ceiling(n_rows * 0.75),   
    row_number = row_number()
  ) %>%
  filter(row_number > cutoff) %>%
  pull(index)
ms2_qtof_25pure <- feature_spectra_centro[keep_idx]
```

### Test the results on the previous feature

```{r, echo = FALSE}
#remove 75%


# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_25pure$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(ms2_qtof_25pure)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)
  )

#Get indices of spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# Run across only the filtered spectra (FT10698 | DDA3)
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  # extract m/z vector (first column)
  mz.data <- peaksData(ms2_qtof_25pure)[[spec_index]][,1]
  spec_label <-  j
  analyze_spectrum(mz.data, formulas, masses, spec_label)
}))


# Filter only ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# Table: how many spectra have N matches
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  
  cat("\nRemove 75%: Spectra count by number of matches (how many spectra had N matches)\n")
  print(table(table(filtered_results$Spectrum)))
}

#remove 50%


# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_50pure$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(ms2_qtof_50pure)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)
  )

#Get indices of spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# Run across only the filtered spectra (FT10698 | DDA3)
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  # extract m/z vector (first column)
  mz.data <- peaksData(ms2_qtof_50pure)[[spec_index]][,1]
  spec_label <- j
  analyze_spectrum(mz.data, formulas, masses, spec_label)
}))


# Filter only ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# Table: how many spectra have N matches
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  
  cat("\nRemove 50%: Spectra count by number of matches (how many spectra had N matches)\n")
  print(table(table(filtered_results$Spectrum)))
}

#remove 30%


# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_70pure$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(ms2_qtof_70pure)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)
  )

#Get indices of spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# Run across only the filtered spectra (FT10698 | DDA3)
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  # extract m/z vector (first column)
  mz.data <- peaksData(ms2_qtof_70pure)[[spec_index]][,1]
  spec_label <- j
  analyze_spectrum(mz.data, formulas, masses, spec_label)
}))


# Filter only ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# Table: how many spectra have N matches
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  
  cat("\nRemove 30%: Spectra count by number of matches (how many spectra had N matches)\n")
  print(table(table(filtered_results$Spectrum)))
}

#remove 20%


# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_80pure$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(ms2_qtof_80pure)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)
  )

#Get indices of spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# Run across only the filtered spectra (FT10698 | DDA3)
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  # extract m/z vector (first column)
  mz.data <- peaksData(ms2_qtof_80pure)[[spec_index]][,1]
  spec_label <- j
  analyze_spectrum(mz.data, formulas, masses, spec_label)
}))


# Filter only ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# Table: how many spectra have N matches
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  
  cat("\nRemove 20%: Spectra count by number of matches (how many spectra had N matches)\n")
  print(table(table(filtered_results$Spectrum)))
}

#remove 10%


# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_90pure$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(ms2_qtof_90pure)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)
  )

#Get indices of spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# Run across only the filtered spectra (FT10698 | DDA3)
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  # extract m/z vector (first column)
  mz.data <- peaksData(ms2_qtof_90pure)[[spec_index]][,1]
  spec_label <- j
  analyze_spectrum(mz.data, formulas, masses, spec_label)
}))


# Filter only ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# Table: how many spectra have N matches
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  
  cat("\nRemove 10%: Spectra count by number of matches (how many spectra had N matches)\n")
  print(table(table(filtered_results$Spectrum)))
}

#remove 5%


# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_95pure$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(ms2_qtof_95pure)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)
  )

#Get indices of spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# Run across only the filtered spectra (FT10698 | DDA3)
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  # extract m/z vector (first column)
  mz.data <- peaksData(ms2_qtof_95pure)[[spec_index]][,1]
  spec_label <- j
  analyze_spectrum(mz.data, formulas, masses, spec_label)
}))


# Filter only ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# Table: how many spectra have N matches
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  
  cat("\nRemove 5%: Spectra count by number of matches (how many spectra had N matches)\n")
  print(table(table(filtered_results$Spectrum)))
}

#original

# Rebuild the same f column for the spectra object
filenames <- feature_spectra_centro$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

spectra_df <- as.data.frame(spectraData(feature_spectra_centro)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda)
  )

#Get indices of spectra for FT10698 | DDA3
which_f <- which(spectra_df$f == "FT10698 | DDA3")

# Run across only the filtered spectra (FT10698 | DDA3)
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  # extract m/z vector (first column)
  mz.data <- peaksData(feature_spectra_centro)[[spec_index]][,1]
  spec_label <- j
  analyze_spectrum(mz.data, formulas, masses, spec_label)
}))


# Filter only ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

# Table: how many spectra have N matches
if (nrow(filtered_results) == 0) {
  cat("No matches within 5 ppm found for FT10698 | DDA3\n")
} else {
  
  cat("\nOriginal: Spectra count by number of matches (how many spectra had N matches)\n")
  print(table(table(filtered_results$Spectrum)))
}
```
### Conclusion PrecursorPurity
Removing the 10% low precursor purity spectra does not affect the best spectra.
The number of spectra with 6, 7 or 8 matches remains unchanged.

The number of poorer spectra, that had less than 6 matches with known fragments, decreased. 
Therefore, removing the 10% spectra with lowest precursor purity efficiently enriches the spectra in high quality spectra mz peaks that match known fragments.
Removing 30% or more of the spectra also removed high quality spectra with high numbers of matches. 

### Discard 10% spectra with lowest precursor purity

```{r}
# extract the spectra data as a dataframe
library(dplyr)
filenames <- feature_spectra$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")



purity_df <- as.data.frame(spectraData(feature_spectra_centro)) %>%
  mutate(
    index = row_number(),
    f = paste(feature_id, "|", dda) 
  )

# 10% lowest
keep_idx <- purity_df %>%
  group_by(f) %>%
  arrange(precursorPurity) %>%
  mutate(
    n_rows = n(),
    cutoff = ceiling(n_rows * 0.10),   
    row_number = row_number()
  ) %>%
  filter(row_number > cutoff) %>%
  pull(index)
ms2_qtof_precpur <- feature_spectra_centro[keep_idx]

#number of spectra:
dim(spectraData(ms2_qtof_precpur))[1]

#number of peaks per spectrum:
counts <- sapply(peaksData(ms2_qtof_precpur), function(x) {
  dim(x)[1]
})
summary(counts)
hist(counts)
```

## Merge spectra within groups feature_id and collision energy
create a single spectrum for each feature per collision energy, keeping
only fragments that are present in minimum `minprop` of spectra.
We define that peaks need to be present in at least two different spectra, otherwise they are considered noise.
peaks = "intersect": keep only peaks in the resulting peak matrix that are present in ⁠>= minProp⁠ proportion of input spectra.
In order to set the value of `minprop` so that it represents at least two spectra, we first need to know how many spectra each group counts.

### Count number of spectra per feature-DDA group
```{r}
# grouping by feature_id and collision energy
# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_precpur$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

f <- paste(ms2_qtof_precpur$feature_id, "|", dda)

#first count spectra per group
groups <- split(seq_along(f), f)
summary(sapply(groups, length))
hist(sapply(groups, length), breaks = 20)


```
Most groups have less than 10 spectra. 
Therefore, for the minimal proportion to be composed of minimum two peaks, we need to set `minprop = 20`.

###check parameters Combine spectra
```{r}
# grouping by feature_id and collision energy
# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_precpur$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

f <- paste(ms2_qtof_precpur$feature_id, "|", dda)


ms2_qtof_ftdda_20 <- combineSpectra(ms2_qtof_precpur, ppm = 5, peaks = "intersect", weighted = TRUE, minProp = 0.2, intensityFun = mean, f = f, p = rep(1L, length(ms2_qtof_precpur)))

#number of spectra:
dim(spectraData(ms2_qtof_ftdda_20))[1]

#number of peaks per spectrum
counts <- sapply(peaksData(ms2_qtof_ftdda_20), function(x) {
  dim(x)[1]
})

summary(counts)

ms2_qtof_ftdda_50 <- combineSpectra(ms2_qtof_precpur, ppm = 5, peaks = "intersect", weighted = TRUE, minProp = 0.50, intensityFun = mean, f = f, p = rep(1L, length(ms2_qtof_precpur)))

#number of spectra:
dim(spectraData(ms2_qtof_ftdda_50))[1]

#number of peaks per spectrum
counts <- sapply(peaksData(ms2_qtof_ftdda_50), function(x) {
  dim(x)[1]
})

summary(counts)
```
The median number of peaks per spectrum went down from 142 to 56 with `minProp = 0.2`and 19 with `minProp = 0.5`.
However, there were also spectra created with many more peaks. Maximum number of peaks increased from 1274 to 3693 with `minProp = 0.2`and 1402 with `minProp = 0.5`.

### check FT10698 with collision energy  DDA3

We first focus on the spectra of the known compound, and check if all peaks matching known fragments are retained.


test the results on the previous feature

```{r}
#minprop 0.20
# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_ftdda_20$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

f <- paste(ms2_qtof_ftdda_20$feature_id, "|", dda)


# 2. Identify spectra for FT10698 | DDA3
which_f <- which(f == "FT10698 | DDA3")

# 3. Run analysis only on these spectra
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  mz.data <- peaksData(ms2_qtof_ftdda_20)[[spec_index]][,1]  # first column = m/z
  analyze_spectrum(mz.data, formulas, masses, j)
}))

# 4. Filter matches with ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

filtered_results

#minprop 0.50
# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_ftdda_50$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

f <- paste(ms2_qtof_ftdda_50$feature_id, "|", dda)


# 2. Identify spectra for FT10698 | DDA3
which_f <- which(f == "FT10698 | DDA3")

# 3. Run analysis only on these spectra
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  mz.data <- peaksData(ms2_qtof_ftdda_50)[[spec_index]][,1]  # first column = m/z
  analyze_spectrum(mz.data, formulas, masses, j)
}))

# 4. Filter matches with ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

filtered_results
```
Both using with `minProp = 0.2`and `minProp = 0.5`, the resulting combined spectrum for feature FT10698 with collision energy  DDA3 contains still all 8 fragments matching known molecular formulas. The combined m/z values for these 8 fragments were exactly the same with the two settings.
For some of these fragments the ppm difference slightly increased, for others the ppm difference decreased. In general, the difference between observed an theoretical mz was below 2 ppm (one exception at 2.4 ppm)
Because with `minProp = 0.5` results overall in lower numbers of peaks per spectrum, we go on with  `minProp = 0.5`.

###Combine spectra
```{r}
# grouping by feature_id and collision energy
# Rebuild the same f column for the spectra object
filenames <- ms2_qtof_precpur$dataOrigin
dda <- str_extract(filenames, "DDA[0-9]")

f <- paste(ms2_qtof_precpur$feature_id, "|", dda)


ms2_qtof_ftdda <- combineSpectra(ms2_qtof_precpur, ppm = 5, peaks = "intersect", weighted = TRUE, minProp = 0.5, intensityFun = mean, f = f, p = rep(1L, length(ms2_qtof_precpur)))

#number of spectra:
dim(spectraData(ms2_qtof_ftdda))[1]

#number of peaks per spectrum
counts <- sapply(peaksData(ms2_qtof_ftdda), function(x) {
  dim(x)[1]
})

summary(counts)


```

## Grouping by feature_id for different collision energies

```{r}
# grouping by feature_id for different collision energies
f = ms2_qtof_ftdda$feature_id
ms2_qtof <- combineSpectra(ms2_qtof_ftdda, ppm = 5, peaks = "union",intensityFun = sum,weighted = TRUE, f = f, p = rep(1L, length(ms2_qtof_ftdda)))

spectraData(ms2_qtof)$spectrum_type <- "merged_MSMS_all_energies"

```

### check FT10698 merged spectrum for all collision energies

```{r}
# Rebuild the same f column for the spectra object
filenames <- ms2_qtof$dataOrigin

f <- ms2_qtof$feature_id


# 2. Identify spectra for FT10698 | DDA3
which_f <- which(f == "FT10698")

# 3. Run analysis only on these spectra
all_results <- do.call(rbind, lapply(seq_along(which_f), function(j) {
  spec_index <- which_f[j]
  mz.data <- peaksData(ms2_qtof)[[spec_index]][,1]  # first column = m/z
  analyze_spectrum(mz.data, formulas, masses, j)
}))

# 4. Filter matches with ppm < 5
filtered_results <- subset(all_results, abs(PPM) < 5)

filtered_results
```
The combined spectrum for feature FT10698 (with collision energy  DDA3)all collision energies) contains still all 8 fragments matching known molecular formulas.
The accuracy furhter increased, and the difference between observed an theoretical mz are now all below 2 ppm.


```{r}
length(unique(unlist(ms2_qtof$feature_id)))

length(unique(spectraData(ms2_qtof)$feature_id))
```

As a result we have `r length(unique(spectraData(ms2_qtof)$feature_id))`
unique feature
### check number of peaks per spectrum

```{r}

#number of peaks per spectrum
counts <- sapply(peaksData(ms2_qtof), function(x) {
  dim(x)[1]
})

summary(counts)

hist(counts, breaks = 100)
hist(counts, breaks = 100, xlim = c(0,400))
```

```{r}

#inspect some spectra that have more than 100 peaks
peak_counts <- sapply(peaksData(ms2_qtof), nrow)
over100 <- which(peak_counts > 100)
idx <- over100[1]
peaksData(ms2_qtof)[[idx]]
spec <- peaksData(ms2_qtof)[[idx]]
o <- order(spec[, 2], decreasing = TRUE)
spec[o,][1:50,]


```

## Keep 100 peaks with highest intensity

```{r}
# Keep only the 100 peaks with the highest intensity
top100 <- function(x, ...) {
    # If fewer than 100 peaks, keep them all
    if (length(x) <= 100) {
        rep(TRUE, length(x))
    } else {
        # Keep TRUE for the 100 highest intensities
        rank(-x, ties.method = "first") <= 100
    }
}

# Apply the filter
ms2_qtof <- filterIntensity(ms2_qtof, intensity = top100)
peak_counts <- sapply(peaksData(ms2_qtof), nrow)
summary(peak_counts)
```



Now we save the object

```{r}
save(ms2_qtof, file = file.path(pth, "data", "ms2_qtof.RData"))
```

## Session information

```{r}
sessionInfo()
```
